
```{r, echo = FALSE}
library(circlize)
```

# High-level genomic functions

In this chapter, several high-level functions which create tracks are introduced.

## Ideograms

`circos.initializeWithIdeogram()` initializes the circular plot and adds ideogram track
if the cytoband data is available. Actually, the ideograms are drawn by `circos.genomicIdeogram()`.
`circos.genomicIdeogram()` creates a thin track of ideograms and can be used anywhere
in the circle. By default it adds ideograms for human genome hg19 (Figure \@ref(fig:ideogram)).

```{r ideogram, fig.cap = "Circular ideograms."}
circos.initializeWithIdeogram(plotType = c("labels", "axis"))
circos.track(ylim = c(0, 1))
circos.genomicIdeogram() # put ideogram as the third track
circos.genomicIdeogram(track.height = 0.2)
```

## Heatmaps

Matrix which corresponds to genomic regions can be visualized as heatmaps. Heatmaps
complete fill the track and there are connection lines connect heatmaps and positions
in the genome. `circos.genomicHeamtap() draws connection lines and heatmaps as two tracks
and combines them as a integrated track.

Generally, all numeric columns (excluding the first three columns) in the input data frame
are used to make the heatmap. Colors can be specfied as a color matrix or a color mapping
function generated by `colorRamp2()`.

The height of the connection line track and the heatmap track can be set by `connection_height`
and `heatmap_height` arguments. Also parameters for the styles of lines and rectangle borders
can be adjusted, please check the help page of `circos.genomicHeatmap()`.

```{r genomic-heatmap1, eval = FALSE}
circos.initializeWithIdeogram()
bed = generateRandomBed(nr = 100, nc = 4)
col_fun = colorRamp2(c(-1, 0, 1), c("green", "black", "red"))
circos.genomicHeatmap(bed, col = col_fun, side = "inside", border = "white")
circos.clear()
```

In the left figure in Figure \@ref(fig:genomic-heatmap), the heatmaps are put inside the 
normal genomic track. Heatmaps are also be put outside the normal genomic track by setting
`side = "outside"` (Figure \@ref(fig:genomic-heatmap), right).

```{r genomic-heatmap2, eval = FALSE}
circos.initializeWithIdeogram(plotType = NULL)
circos.genomicHeatmap(bed, col = col_fun, side = "outside",
    line_col = as.numeric(factor(bed[[1]])))
circos.genomicIdeogram()
circos.clear()
```

```{r genomic-heatmap, echo = FALSE, fig.width = 8, fig.height = 4, fig.cap = "Genomic heamtaps."}
chunks <- knitr:::knit_code$get()
par(mfrow = c(1, 2))
eval(parse(text = chunks[["genomic-heatmap1"]]))
eval(parse(text = chunks[["genomic-heatmap2"]]))
```

## Labels

`circos.genomicLabels()` adds text labels for regions specified.
Positions of labels are automatically adjusted so that they do not
overlap to each other.

Similar as `circos.genomicHeatmap()`, `circos.genomicLabels()` also 
creates two tracks where one for the connection lines and one for the 
labels. You can set the height of the labels track to be the maximum
width of labels by `labels_track = max(strwidth(labels))`. `padding`
argument controls the gap between two neighbouring labels.

```{r genomic-labels1, eval = FALSE}
circos.initializeWithIdeogram()
bed = generateRandomBed(nr = 50, fun = function(k) sample(letters, k, replace = TRUE))
bed[1, 4] = "aaaaa"
circos.genomicLabels(bed, labels.column = 4, side = "inside")
circos.clear()
```

Similarlly, labels can be put outside of the normal genomic track (Figure \@ref(fig:genomic-labels) right).

```{r genomic-labels2, eval = FALSE}
circos.initializeWithIdeogram(plotType = NULL)
circos.genomicLabels(bed, labels.column = 4, side = "outside",
	col = as.numeric(factor(bed[[1]])), line_col = as.numeric(factor(bed[[1]])))
circos.genomicIdeogram()
circos.clear()
```

```{r genomic-labels, echo = FALSE, fig.width = 8, fig.height = 4, fig.cap = "Genomic labels."}
chunks <- knitr:::knit_code$get()
par(mfrow = c(1, 2))
eval(parse(text = chunks[["genomic-labels1"]]))
eval(parse(text = chunks[["genomic-labels2"]]))
```

## Genomic density and Rainfall plot

Rainfall plots are used to visualize the distribution of genomic regions in the genome.
Rainfall plots are particularly useful to identify clusters of regions. In the
rainfall plot, each dot represents a region. The x-axis corresponds to the
genomic coordinate, and the y-axis corresponds to the minimal distance (log10
transformed) of the region to its two neighbouring regions. A cluster of regions will
appear as a “rainfall” in the plot.

`circos.genomicRainfall()` calculates neighbouring distance for each region
and draw as points on the plot. Since `circos.genomicRainfall()` generates data on
y-direction (`log10(distance)`), it is actually a high-level function which
creates a new track.

The input data can be a single data frame or a list of data frames.

```{r, eval = FALSE}
circos.genoimcRainfall(bed)
circos.genoimcRainfall(bed_list, col = c("red", "green"))
```

However, if the amount of regions in a cluster is high, dots will overlap, and
direct assessment of the number and density of regions in the cluster will be
impossible. To overcome this limitation, additional tracks are added which
visualize the genomic density of regions (defined as the fraction of a genomic
window that is covered by genomic regions).

`circos.genomicDensity()` calculates how much a genomic window is covered by
regions in `bed`. It is also a high-level function and creates a new track.

The input data can be a single data frame or a list of data frames.

```{r, eval = FALSE}
circos.genomicDensity(bed)
circos.genomicDensity(bed, baseline = 0)
circos.genomicDensity(bed, window.size = 1e6)
circos.genomicDensity(bedlist, col = c("#FF000080", "#0000FF80"))
```

Following example makes a rainfall plot for differentially methylated regions
(DMR) and their genomic densities. In the plot, red corresponds to hyper-methylated 
DMRs (gain of methylation) and blue corresponds to hypo-methylated
DMRs (loss of methylation). You may see how the combination of rainfall track
and genomic density track helps to get a more precise inference of the
distribution of DMRs on genome (Figure \@ref(fig:genomic-rainfall)).

```{r genomic-rainfall, fig.width = 6, fig.height = 6, fig.cap = "Genomic rainfall plot and densities."}
load(system.file(package = "circlize", "extdata", "DMR.RData"))
circos.initializeWithIdeogram(chromosome.index = paste0("chr", 1:22))

bed_list = list(DMR_hyper, DMR_hypo)
circos.genomicRainfall(bed_list, pch = 16, cex = 0.4, col = c("#FF000080", "#0000FF80"))
circos.genomicDensity(DMR_hyper, col = c("#FF000080"), track.height = 0.1)
circos.genomicDensity(DMR_hypo, col = c("#0000FF80"), track.height = 0.1)
circos.clear()
```
