```{r, echo = FALSE}
library(circlize)
```

# Revisit zooming

Zooming

After we have learned the basic and advanced technique of **circlize**

Circular visualization of DMRs from tagmentation-based WGBS
======================================================

[Tagmentation-based whole-genome bisulfite sequencing](http://www.nature.com/nprot/journal/v8/n10/full/nprot.2013.118.html) 
(T-WGBS) is a technology which can examine only a minor fraction of methylome of interest.
Circular plot can be used to [visualize genome-wide distribution of differentially methylation
regions (DMRs)](http://jokergoo.github.io/circlize/example/railfallplot.html). 
In this post, we demonstrate how to visualize DMRs which are detected from T-WGBS data in a circular plot by **circlize** package.

In [`tagments_WGBS_DMR.RData`](../../data/tagments_WGBS_DMR.RData), `tagments` contains regions which are sequenced, 
`DMR1` and `DMR2` contain DMRs for two patients detectd in tagment regions. Correspondance between tagment regions
and DMRs can be checked by row names of `tagments` and `tagment` column in `DMR1` or `DMR2`.

```{r}
load("../data/tagments_WGBS_DMR.RData")
tagments[1:2, ]
DMR1[1:2, ]
DMR2[1:2, ]
```

Chromosomes (e.g. `chr1`, `chr2`) and tagments (e.g. `chr1-44876009-45016546`, `chr1-90460304-90761641`)
are actually different types of categories and **circlize** can only deal with one type at a time.
In order to merge chromosomes and tagment regions into one same plot, the strategy is to create two independent
circular plot but overlay the second one directly to the first one by specifying `par(new = TRUE)`.

First we draw the ideograms as well as the chromosome names and we call it 'the first circular plot'.

```{r first_plot_1, eval = FALSE}
library(circlize)

circos.par(gap.degree = 2, start.degree = 90)
circos.initializeWithIdeogram(chromosome.index = paste0("chr", 1:22), plotType = c("ideogram", "labels"),
	ideogram.height = 0.03)
```

```{r, echo = FALSE}
chunks <- knitr:::knit_code$get()
eval(parse(text = chunks[["first_plot_1"]]))
```

To make the correspondance between two circular plots, we need mapping of a same genomic position
between two circular plots. Here the key solution is the polar coordinate system because all circular
plot created by **circlize** are in a same polar coordinate system. With using `circlize()` function, we can transform
genomic positions in one circular plot into polar coordinates and we can apply `reverse.circlize()` to get 
the new genomic positions in the second circualr plot.

```{r first_plot_2, eval = FALSE}
# calculate position of each tagment measured in the polar coordinate system
for(i in seq_len(nrow(tagments))) {
	tagments[i, "theta1"] = circlize(tagments[i, 2], 1, sector.index = tagments[i, 1])[1, 1]
	tagments[i, "theta2"] = circlize(tagments[i, 3], 1, sector.index = tagments[i, 1])[1, 1]
}
```

Since the second circular plot which visualizes DMRs will be put inside the ideograms, we need to know
the radius which inside the ideogram. Since the ideogram in the first circular plot is the last track, we can
use `circlize:::get_most_inside_radius()` to get the position.

```{r, first_plot_3, eval = FALSE}
r = circlize:::get_most_inside_radius()
circos.clear()
```

```{r, echo = FALSE}
eval(parse(text = chunks[["first_plot_2"]]))
eval(parse(text = chunks[["first_plot_3"]]))
```

Now we make the second circular plot which visualizes DMRs. Here `par(new = TRUE)` is set to directly overlay
to the first plot. 

in `circos.par()`, `start.degree` can be adjusted to rotate the second circular plot to make better correspondance
between two circular plots.

`chr_bg_color` is defined here to enhance discrimination of different chromosomes.

```{r second_plot_1, eval = FALSE}
par(new = TRUE)

set.seed(123)
chr_bg_color = rand_color(22, transparency = 0.8)
names(chr_bg_color) = paste0("chr", 1:22)

circos.par(cell.padding = c(0.02, 0, 0.02, 0), gap.degree = c(rep(1, nrow(tagments)-1), 10), 
	start.degree = 75, points.overflow.warning = FALSE)
circos.initialize(factors = factor(rownames(tagments), levels = rownames(tagments)), 
	xlim = as.matrix(tagments[, 2:3]))
```

In the first track of the second circular plot, we need to draw kind of connections between two circular
plots. Since we know the coordinates of tagment regions in the polar coordinate system, here we use 
`reverse.circlize()` to transform back to the corresponding genomic positions (or coordinate in the data coordinate system).

Since the second plot is drawn inside the ideograms, `r` is set as the outter margin of the second plot.


```{r second_plot_2, eval = FALSE}
circos.track(ylim = c(0, 1), panel.fun = function(x, y) {
	
	si = get.cell.meta.data("sector.index")
	chr = gsub("^(chr.*?)-.*$", "\\1", si)
	theta1 = tagments[si, "theta1"]
	theta2 = tagments[si, "theta2"]
	
	xlim = get.cell.meta.data("cell.xlim")
	ylim = get.cell.meta.data("cell.ylim")

	cell.top.radius = get.cell.meta.data("cell.top.radius")
	
	# map from polar coordinate system to data coordinate system
	df = reverse.circlize(c(theta1, theta2), c(cell.top.radius, cell.top.radius))

	x21 = df[1, 1]
	x22 = df[2, 1]
	y21 = df[1, 2]
	y22 = df[2, 2]
	x11 = xlim[1]
	x12 = xlim[2]
	y11 = ylim[1]
	y12 = ylim[1]
	circos.polygon(c(x11, x11, x21, x21, x22, x22, x12, x12, x11),
		           c(y11, (y21 - y11)/3, (y21 - y11)/3*2, y21, y22, (y22 - y12)/3*2, (y22 - y12)/3, y12, y11), 
		           col = chr_bg_color[chr])

}, track.margin = c(0, 1 - r), cell.padding = c(0, 0, 0, 0), bg.border = NA, track.height = 0.1)
```

```{r, echo = FALSE}
eval(parse(text = chunks[["first_plot_1"]]))
eval(parse(text = chunks[["first_plot_2"]]))
eval(parse(text = chunks[["first_plot_3"]]))
eval(parse(text = chunks[["second_plot_1"]]))
eval(parse(text = chunks[["second_plot_2"]]))
circos.clear()
```

Now we add DMR tracks. In each cell (plotting region), we added the background with colors,
reference lines and points.

```{r DMR1, eval = FALSE}
max_abs = max(abs(c(DMR1$methDiff, DMR2$methDiff)))
max_abs = ceiling(max_abs*10)/10
circos.track(ylim = c(-max_abs, max_abs), panel.fun = function(x, y) {
	si = get.cell.meta.data("sector.index")
	chr = gsub("^(chr\\d+).*$", "\\1", si)
	xlim = get.cell.meta.data("cell.xlim")
	ylim = get.cell.meta.data("cell.ylim")

	circos.rect(xlim[1], ylim[1], xlim[2], ylim[2], col = chr_bg_color[[chr]])
	for(h in seq(-max_abs, max_abs, by = 0.3)) {
		circos.lines(xlim, c(h, h), lty = 3, col = "#AAAAAA")
	}

	circos.lines(xlim, c(0, 0), lty = 3, col = "#888888")

	subset = DMR1[DMR1$tagment == si, , drop = FALSE]
	if(nrow(subset) > 0) {
		circos.points((subset[[2]] + subset[[3]])/2, subset$methDiff, 
			col = ifelse(subset$methDiff > 0, "#E41A1C", "#377EB8"), pch = 16, cex = 0.5)
	}

}, bg.border = 1, track.height = 0.15)
```

Also we add y-axis and labels on the left of the first tagment.

```{r DMR1_axis, eval = FALSE}
first_sector = get.all.sector.index()[1]
circos.yaxis(side = "left", at = seq(-0.6, 0.6, by = 0.3), sector.index = first_sector,
	labels.cex = 0.4)
xlim = get.cell.meta.data("cell.xlim", sector.index = first_sector)
ylim = get.cell.meta.data("cell.ylim", sector.index = first_sector)
circos.text(xlim[1], mean(ylim), "d1", facing = "clockwise", niceFacing = TRUE, cex = 0.8, 
	adj = c(0.5, degree(6)), sector.index = first_sector)
```

```{r, echo = FALSE}
eval(parse(text = chunks[["first_plot_1"]]))
eval(parse(text = chunks[["first_plot_2"]]))
eval(parse(text = chunks[["first_plot_3"]]))
eval(parse(text = chunks[["second_plot_1"]]))
eval(parse(text = chunks[["second_plot_2"]]))
eval(parse(text = chunks[["DMR1"]]))
eval(parse(text = chunks[["DMR1_axis"]]))
circos.clear()
```

Add the DMRs for the second patient.

```{r DMR2, eval = FALSE}
circos.track(ylim = c(-max_abs, max_abs), panel.fun = function(x, y) {
	si = get.cell.meta.data("sector.index")
	chr = gsub("^(chr\\d+).*$", "\\1", si)
	xlim = get.cell.meta.data("cell.xlim")
	ylim = get.cell.meta.data("cell.ylim")
	circos.rect(xlim[1], ylim[1], xlim[2], ylim[2], col = chr_bg_color[[chr]])
	for(h in seq(-max_abs, max_abs, by = 0.3)) {
		circos.lines(xlim, c(h, h), lty = 3, col = "#AAAAAA")
	}
	circos.lines(xlim, c(0, 0), lty = 3, col = "#888888")
	subset = DMR2[DMR2$tagment == si, , drop = FALSE]
	if(nrow(subset) > 0) {
		circos.points((subset[[2]] + subset[[3]])/2, subset$methDiff, 
			col = ifelse(subset$methDiff > 0, "#E41A1C", "#377EB8"), pch = 16, cex = 0.5)
	}
}, bg.border = 1, track.height = 0.15)
circos.yaxis(side = "left", at = seq(-0.6, 0.6, by = 0.3), sector.index = first_sector,
	labels.cex = 0.4)
xlim = get.cell.meta.data("cell.xlim", sector.index = first_sector)
ylim = get.cell.meta.data("cell.ylim", sector.index = first_sector)
circos.text(xlim[1], mean(ylim), "d2", facing = "clockwise", niceFacing = TRUE, cex = 0.8, 
	adj = c(0.5, degree(6)), sector.index = first_sector)
```

For the most inside track, we explicitely show the color of chromosomes to make it easier to correspond
between target regions to chromosomes.

```{r last, eval = FALSE}
circos.track(ylim = c(0, 1), panel.fun = function(x, y) {
	cate = get.cell.meta.data("sector.index")
	chr = gsub("^(chr\\d+).*$", "\\1", cate)
	xlim = get.cell.meta.data("cell.xlim")
	ylim = get.cell.meta.data("cell.ylim")
	circos.rect(xlim[1], ylim[1], xlim[2], ylim[2], col = gsub("\\d\\d$", "", chr_bg_color[[chr]]))
}, track.height = 0.02, cell.padding = c(0, 0, 0, 0))

circos.clear()

legend("center", pch = 16, legend = c("Hyper-DMR", "Hypo-DMR"), col = c("#E41A1C", "#377EB8"))

par(new = FALSE)
```

```{r, echo = FALSE}
eval(parse(text = chunks[["first_plot_1"]]))
eval(parse(text = chunks[["first_plot_2"]]))
eval(parse(text = chunks[["first_plot_3"]]))
eval(parse(text = chunks[["second_plot_1"]]))
eval(parse(text = chunks[["second_plot_2"]]))
eval(parse(text = chunks[["DMR1"]]))
eval(parse(text = chunks[["DMR1_axis"]]))
eval(parse(text = chunks[["DMR2"]]))
eval(parse(text = chunks[["last"]]))
circos.clear()
```

<!-- uid=segmentation_WGBS_dmr -->