```{r, echo = FALSE}
set.seed(999)
mat = matrix(sample(18, 18), 3, 6) 
rownames(mat) = paste0("S", 1:3)
colnames(mat) = paste0("E", 1:6)
grid.col = c(S1 = "red", S2 = "green", S3 = "blue",
    E1 = "grey", E2 = "grey", E3 = "grey", E4 = "grey", E5 = "grey", E6 = "grey")
library(circlize)
```

# Advanced usage of `chordDiagram()`

The default style of `chordDiagram()` is somehow enough for most visualization
tasks, still you can have more configurations on the plot.

The usage is same for both ajacency matrix and ajacency list, so we only
demonstrate with the matrix.

## Organization of tracks

By default, `chordDiagram()` creates two tracks, one track for labels and one
track for grids with axes.

```{r, echo = 2:3}
pdf(NULL)
chordDiagram(mat)
circos.info()
invisible(dev.off())
```

These two tracks can be controlled by `annotationTrack` argument. Available
values for this argument are `grid`, `name` and `axis`. The height of
annotation tracks can be set by `annotationTrackHeight` which is the
percentage to the radius of unit circle and can be set by `uh()` function with
an absolute unit. Axes are only added if `grid` is set in `annotationTrack` 
(Figure \@ref(fig:chord-diagram-default-track)).

```{r chord-diagram-default-track, echo = -1, fig.width = 8, fig.height = 8/3, fig.cap = "Track organization in `chordDiagram()`."}
par(mfrow = c(1, 3))
chordDiagram(mat, grid.col = grid.col, annotationTrack = "grid")
chordDiagram(mat, grid.col = grid.col, annotationTrack = c("name", "grid"),
    annotationTrackHeight = c(0.03, 0.01))
chordDiagram(mat, grid.col = grid.col, annotationTrack = NULL)
```

Several empty tracks can be allocated before Chord diagram is drawn. Then self-defined graphics can
be added to these empty tracks afterwards. The number of pre-allocated tracks can be set 
through `preAllocateTracks`.

```{r echo = 2:3}
pdf(NULL)
chordDiagram(mat, preAllocateTracks = 2)
circos.info()
invisible(dev.off())
```

The default settings for pre-allocated tracks are:

```{r, eval = FALSE}
list(ylim = c(0, 1),
     track.height = circos.par("track.height"),
     bg.col = NA,
     bg.border = NA,
     bg.lty = par("lty"),
     bg.lwd = par("lwd"))
```

The default settings for pre-allocated tracks can be overwritten by specifying `preAllocateTracks`
as a list.

```{r, eval = FALSE}
chordDiagram(mat, annotationTrack = NULL,
    preAllocateTracks = list(track.height = 0.3))
```

If more than one tracks need to be pre-allocated, just specify `preAllocateTracks`
as a list which contains settings for each track:

```{r, eval = FALSE}
chordDiagram(mat, annotationTrack = NULL,
    preAllocateTracks = list(list(track.height = 0.1),
                             list(bg.border = "black")))
```

By default `chordDiagram()` provides poor supports for customization of sector
labels and axes, but with `preAllocateTracks` it is rather easy to customize
them. Such customization will be introduced in next section.

## Customize sector labels

In `chordDiagram()`, there is no argument to control the style of sector
labels, but this can be done by first pre-allocating an empty track and
customizing the labels in it later. In the following example, one track is
firstly allocated and a Chord diagram is added without label track and axes.
Later, the first track is updated with adding labels with clockwise facings
(Figure \@ref(fig:chord-diagram-labels-show)).

```{r chord-diagram-labels-show, fig.cap = "Change label directions."}
chordDiagram(mat, grid.col = grid.col, annotationTrack = "grid", 
    preAllocateTracks = list(track.height = max(strwidth(unlist(dimnames(mat))))))
# we go back to the first track and customize sector labels
circos.trackPlotRegion(track.index = 1, panel.fun = function(x, y) {
    circos.text(CELL_META$xcenter, CELL_META$ylim[1], CELL_META$sector.index, 
        facing = "clockwise", niceFacing = TRUE, adj = c(0, 0.5))
}, bg.border = NA) # here set bg.border to NA is important
```

In the following example, the labels are put on the grids (Figure \@ref(fig:chord-diagram-labels-inside)). 
Please note `circos.text()` and
`get.cell.meta.data()` can be used outside `panel.fun` if the sector index and
track index are specified explicitly.

```{r chord-diagram-labels-inside, fig.cap = "Put sector labels to the grid."}
chordDiagram(mat, grid.col = grid.col, 
    annotationTrack = c("grid", "axis"), annotationTrackHeight = uh(5, "mm"))
for(si in get.all.sector.index()) {
    xlim = get.cell.meta.data("xlim", sector.index = si, track.index = 1)
    ylim = get.cell.meta.data("ylim", sector.index = si, track.index = 1)
    circos.text(mean(xlim), mean(ylim), si, sector.index = si, track.index = 1, 
        facing = "bending.inside", niceFacing = TRUE, col = "white")
}
```

For the last example in this section, if the width of the sector is less than
20 degree, the labels are added in the radical direction (Figure \@ref(fig:chord-diagram-labels-multile-style)).

```{r chord-diagram-labels-multile-style, fig.cap = "Adjust label direction according to the width of sectors."}
set.seed(123)
mat2 = matrix(rnorm(100), 10)
chordDiagram(mat2, annotationTrack = "grid", 
    preAllocateTracks = list(track.height = max(strwidth(unlist(dimnames(mat))))))
circos.trackPlotRegion(track.index = 1, panel.fun = function(x, y) {
    xlim = get.cell.meta.data("xlim")
    xplot = get.cell.meta.data("xplot")
    ylim = get.cell.meta.data("ylim")
    sector.name = get.cell.meta.data("sector.index")

    if(abs(xplot[2] - xplot[1]) < 20) {
        circos.text(mean(xlim), ylim[1], sector.name, facing = "clockwise",
            niceFacing = TRUE, adj = c(0, 0.5), col = "red")
    } else {
        circos.text(mean(xlim), ylim[1], sector.name, facing = "inside", 
            niceFacing = TRUE, adj = c(0.5, 0), col= "blue")
    }
}, bg.border = NA)
```

When you set direction of sector labels as radical (`clockwise` or
`reverse.clockwise`), if the labels are too long and exceed your figure
region, you can either decrease the size of the font or set `canvas.xlim` and
`canvas.ylim` parameters in `circos.par()` to wider intervals.

## Customize sector axes

Axes are helpful to correspond to the absolute values of links. By default
`chordDiagram()` adds axes on the grid track. But it is easy to customize one
with self-defined code.

In following example code, we draw another type of axes which show relative
percent on sectors. We first pre-allocate an empty track by
`preAllocateTracks` and come back to this track to add axes later.

You may see we add the first axes to the top of second track. You can also
put them to the bottom of the first track.

```{r chord_diagram_axes_two, eval = FALSE}
# similar as the previous example, but we only plot the grid track
chordDiagram(mat, grid.col = grid.col, annotationTrack = "grid", 
    preAllocateTracks = list(track.height = uh(5, "mm")))
for(si in get.all.sector.index()) {
    circos.axis(h = "top", labels.cex = 0.3, sector.index = si, track.index = 2)
}
```

Now we go back to the first track to add the second type of axes and sector names.
In `panel.fun`, if the sector is less than 30 degree, the break for the axis is set to 0.5
(Figure \@ref(fig:chord-diagram-axes)).


```{r chord_diagram_axes_two2, eval = FALSE}
# the second axis as well as the sector labels are added in this track
circos.trackPlotRegion(track.index = 1, panel.fun = function(x, y) {
    xlim = get.cell.meta.data("xlim")
    ylim = get.cell.meta.data("ylim")
    sector.name = get.cell.meta.data("sector.index")
    xplot = get.cell.meta.data("xplot")
    
    circos.lines(xlim, c(mean(ylim), mean(ylim)), lty = 3) # dotted line
    by = ifelse(abs(xplot[2] - xplot[1]) > 30, 0.2, 0.5)
    for(p in seq(by, 1, by = by)) {
        circos.text(p*(xlim[2] - xlim[1]) + xlim[1], mean(ylim) + 0.1, 
            paste0(p*100, "%"), cex = 0.3, adj = c(0.5, 0), niceFacing = TRUE)
    }
    
    circos.text(mean(xlim), 1, sector.name, niceFacing = TRUE, adj = c(0.5, 0))
}, bg.border = NA)
circos.clear()
```

```{r chord-diagram-axes, echo = FALSE, fig.width = 6, fig.height = 6, fig.cap = "Customize sector axes for Chord diagram."}
chunks <- knitr:::knit_code$get()
eval(parse(text = chunks[["chord_diagram_axes_two"]]))
eval(parse(text = chunks[["chord_diagram_axes_two2"]]))
```

## Put horizontally or vertically symmetric

In Chord diagram, when there are two groups (which correspond to rows and columns
if the input is an adjacency matrix), it is always visually beautiful to rotate the diagram
to be symmetric on horizontal direction or vertical direction. Actually it is quite easy
to calculate a proper degree that needs to be rotated for the circle. 

In the Chord diagram, the total width of row sectors corresponds to the sum of
row sum of the matrix with absolute values and so is for the column sectors.

```{r}
row_sum = sum(rowSums(abs(mat)))
col_sum = sum(colSums(abs(mat)))
```

Assume small gaps between sectors are set to 1 degree and big gaps between row and
column sectors are set to 20 degree.

```{r}
small_gap = 1
big_gap = 20
```

In the circle, there are regions which are covered by small gaps, big gaps and sectors. Since the width of
sectors are proportional to the row sums and/or column sums of the matrix, it is easy to calculate how much
degrees are hold by the row sectors:

```{r}
nr = nrow(mat)
nc = ncol(mat)
n_sector = nr + nc
row_sector_degree = (360 - small_gap*(n_sector - 2) - big_gap*2) * (row_sum/(row_sum + col_sum)) + 
                    small_gap*(nr-1)
```

If the row sectors are put in the right of the circle, we can calculate the "start degree" for the circle.
Note `chordDiagram()` always draw row sectors first and by default the circle goes clock-wisely.

```{r}
start_degree = 90 - (180 - row_sector_degree)/2
```

When there are small gaps and big gaps between sectors, the `gap.after` in
`circos.par()` should be set as a vector. We also added a vertical line which
assists to see the symmetry (Figure \@ref(fig:chord-diagram-sym) left).

```{r chord_diagram_symmetric_1, echo = -1, eval = FALSE}
circos.clear()
gaps = c(rep(small_gap, nrow(mat) - 1), big_gap, rep(small_gap, ncol(mat) - 1), big_gap)
circos.par(gap.after = gaps, start.degree = start_degree)
chordDiagram(mat, grid.col = grid.col)
circos.clear()
abline(v = 0, lty = 2, col = "#00000080")
```

Similarly we can adjust the "start degree" to let the circle looks horizonally
symmetric (Figure \@ref(fig:chord-diagram-sym) right).

```{r chord_diagram_symmetric_2, eval = FALSE}
start_degree = 0 - (180 - row_sector_degree)/2
gaps = c(rep(small_gap, nrow(mat) - 1), big_gap, rep(small_gap, ncol(mat) - 1), big_gap)
circos.par(gap.after = gaps, start.degree = start_degree)
chordDiagram(mat, grid.col = grid.col)
circos.clear()
abline(h = 0, lty = 2, col = "#00000080")
```

```{r chord-diagram-sym, echo = FALSE, fig.width = 8, fig.height = 4, fig.cap = "Rotate Chord diagram."}
par(mfrow = c(1, 2))
chunks <- knitr:::knit_code$get()
eval(parse(text = chunks[["chord_diagram_symmetric_1"]]))
eval(parse(text = chunks[["chord_diagram_symmetric_2"]]))
```

## Compare two Chord diagrams

Normally, in Chord diagram, values in `mat` are normalized to the summation of
the absolute values in the matrix, which means the width for links only
represents relative values. Then, when comparing two Chord diagrams, it is
necessary that unit width of links in the two plots should be represented in a
same scale. This problem can be solved by adding more blank gaps to the Chord
diagram which has smaller matrix.

First, let's plot a Chord diagram. In this Chord diagram, we set larger gaps between rows
and columns for better visualization. Axis on the grid illustrates scale of the values.

```{r chord_diagram_compare_1, eval = FALSE, echo = -3}
mat1 = matrix(sample(20, 25, replace = TRUE), 5)
gap.after = c(rep(2, 4), 10, rep(2, 4), 10)
circos.clear()
circos.par(gap.after = gap.after, start.degree = -10/2)
chordDiagram(mat1, directional = 1, grid.col = rep(1:5, 2))
circos.clear()
```

The second matrix only has half the values in `mat1`.

```{r chord_diagram_compare_2, eval = FALSE}
mat2 = mat1 / 2
```

If the second Chord diagram is plotted in the way as the first one, the two
diagrams will looks exactly the same (except the axes) which makes the
comparison impossible. However, we can adjust the gaps between sectors to make
the scale of the two plots same.

First we calculate the percentage of `mat2` in `mat1`. And then we calculate
the degree which corresponds to the difference. In the following code, `360 -
sum(gap.after)` is the total degree for values in `mat1` (excluding the gaps)
and `blank.degree` corresponds the difference between `mat1` and `mat2`.

```{r chord_diagram_compare_3, eval = FALSE}
percent = sum(abs(mat2)) / sum(abs(mat1))
blank.degree = (360 - sum(gap.after)) * (1 - percent)
```

Since now we have the additional blank gap, we can set it to `circos.par()`
and plot the second Chord Diagram.

```{r chord_diagram_compare_4, eval = FALSE}
big.gap = (blank.degree - sum(rep(2, 8)))/2
gap.after = c(rep(2, 4), big.gap, rep(2, 4), big.gap)
circos.par(gap.after = gap.after, start.degree = -big.gap/2)
chordDiagram(mat2, directional = 1, grid.col = rep(1:5, 2), transparency = 0.5)
circos.clear()
```

Now the scale of the two Chord diagrams (Figure \@ref(fig:chord-diagram-compare)) are the 
same if you compare the scale of axes in the two diagrams.

```{r chord-diagram-compare, echo = FALSE, fig.width = 8, fig.height = 4, fig.cap = "Compare two Chord Diagrams in a same scale."}
par(mfrow = c(1, 2))
eval(parse(text = chunks[["chord_diagram_compare_1"]]))
eval(parse(text = chunks[["chord_diagram_compare_2"]]))
eval(parse(text = chunks[["chord_diagram_compare_3"]]))
eval(parse(text = chunks[["chord_diagram_compare_4"]]))
```

## Multiple-group Chord diagram

Generally `chordDiagram()` function visualizes relations between two groups
(i.e. from rows to columns if the input is an adjacency matrix or from column
1 to column 2 if the input is an adjacency list), however, for
`chordDiagram()`, it actually doesn't need any grouping information. The
visual effect of grouping is just enhanced by setting different gap degrees.
In this case, it is easy to make a Chord diagram with more than two groups.

First let's generate three matrix which contain pairwise relations from three groups:

```{r}
options(digits = 2)
mat1 = matrix(rnorm(25), nrow = 5)
rownames(mat1) = paste0("A", 1:5)
colnames(mat1) = paste0("B", 1:5)

mat2 = matrix(rnorm(25), nrow = 5)
rownames(mat2) = paste0("A", 1:5)
colnames(mat2) = paste0("C", 1:5)

mat3 = matrix(rnorm(25), nrow = 5)
rownames(mat3) = paste0("B", 1:5)
colnames(mat3) = paste0("C", 1:5)
```

Since `chordDiagram()` only accepts one single matrix, here the three
matrix are merged into one big matrix.

```{r}
mat = matrix(0, nrow = 10, ncol = 10)
rownames(mat) = c(rownames(mat2), rownames(mat3))
colnames(mat) = c(colnames(mat1), colnames(mat2))
mat[rownames(mat1), colnames(mat1)] = mat1
mat[rownames(mat2), colnames(mat2)] = mat2
mat[rownames(mat3), colnames(mat3)] = mat3
mat
```

When making the chord diagram, we set larger gaps between groups to identify different groups.
Here we manually adjust `gap.after` in `circos.par()`.

Also we add an additional track in which we add lines to enhance the visual effect of different groups.

```{r}
library(circlize)
circos.par(gap.after = rep(c(rep(1, 4), 8), 3))
chordDiagram(mat, annotationTrack = c("grid", "axis"),
    preAllocateTracks = list(
        track.height = uh(4, "mm"),
        track.margin = c(uh(4, "mm"), 0)
))
circos.track(track.index = 2, panel.fun = function(x, y) {
    sector.index = get.cell.meta.data("sector.index")
    xlim = get.cell.meta.data("xlim")
    ylim = get.cell.meta.data("ylim")
    circos.text(mean(xlim), mean(ylim), sector.index, cex = 0.6, niceFacing = TRUE)
}, bg.border = NA)

highlight.sector(rownames(mat1), track.index = 1, col = "red", 
    text = "A", cex = 0.8, text.col = "white", niceFacing = TRUE)
highlight.sector(colnames(mat1), track.index = 1, col = "green", 
    text = "B", cex = 0.8, text.col = "white", niceFacing = TRUE)
highlight.sector(colnames(mat2), track.index = 1, col = "blue", 
    text = "C", cex = 0.8, text.col = "white", niceFacing = TRUE)
circos.clear()
```

If row names and column names in the big matrix are not grouped, the sector order
can be manually adjusted by `order` argument.

```{r, eval = FALSE}
chordDiagram(mat, order = c(paset0("A", 1:5), paset0("B", 1:5), paset0("C", 1:5)))
```

It is similar way to construct a multiple-group Chord diagram with data frame as input.

```{r, eval = FALSE}
library(reshape2)
df2 = do.call("rbind", list(melt(mat1), melt(mat2), melt(mat3)))
chordDiagram(df2, order = c(paste0("A", 1:5), paste0("B", 1:5), paste0("C", 1:5)))
```
