
# (PART) Visualize Relations {-} 

One unique feature of circular layout is the circular visualization of
relations between objects by links. See examples in
http://circos.ca/intro/tabular_visualization/. The name of such plot is called
[Chord diagram](http://en.wikipedia.org/wiki/Chord_diagram). In **circlize**,
it is easy to plot Chord diagram in a straightforward or customized way.

There are two data formats that represent relations, either adjacency matrix
or adjacency list. In adjacency matrix, value in $i^{th}$ row and $j^{th}$
column represents the relation from object in the $i^{th}$ row and the object
in the $j^{th}$ column where the absolute value measures the strength of the
relation. In adjacency list, relations are represented as a three-column data
frame in which relations come from the first column and to the second column,
and the third column represents the strength of the relation.

Following code shows an example of an adjacency matrix.

```{r}
mat = matrix(1:9, 3)
rownames(mat) = letters[1:3]
colnames(mat) = LETTERS[1:3]
mat
```

And the code in below is an example of a adjacency list.

```{r}
df = data.frame(from = letters[1:3], to = LETTERS[1:3], value = 1:3)
df
```

Actually, it is not difficult to convert between these two formats. There are
also R packages and functions do the conversion such as in **reshape2**
package, `melt()` converts a matrix to a data frame and `dcast()` converts the
data frame back to the matrix.

Chord diagram shows the information of the relation from several levels. 1.
the links are straightforward to show the relations between objects; 2. width
of links are proportional to the strength of the relation which is more
illustrative than other graphic mappings; 3. colors of links can be another
graphic mapping for categorize relations; 4. width of sectors represents total
strength for a link which connects to other objects or is connected from other
objects. You can find an interesting example of using Chord diagram to visualize
leagues system of players clubs by their national team from https://gjabel.wordpress.com/2014/06/05/world-cup-players-representation-by-league-system/ and corresponding code is at
http://jokergoo.github.io/circlize/example/wc2014.html.


# The `chordDiagram()` function

In **circlize** package, there is a `chordDiagram()` function that supports
both adjacency matrix and adjacency list. For different formats of input, the
corresponding format of graphic parameters will be different either. E.g. when
the input is a matrix, since information of the links in the Chord diagram is
stored in the matrix, corresponding graphics for the links sometimes should
also be specified as a matrix, while if the input is a data frame, the graphic
parameters for links only need to be specified as an additional column to the
data frame. However, in many cases, adjacency matrix is directly generated
from upstream analysis and converting it into a adjacency list does not make
sense, e.g. converting a correlation matrix to a adjacency list is obviously a
bad idea. Thus, in this chapter, we will show usage for both adjacency matrix and
list. 

Since the usage for the two types of inputs are highly similar, in this chapter,
we mainly show figures generated from matrix, but still keep the code which uses
adjacency list runable.

## Basic usage

First let's generate a random matrix and the corresponding adjacency list.

```{r}
set.seed(999)
mat = matrix(sample(18, 18), 3, 6) 
rownames(mat) = paste0("S", 1:3)
colnames(mat) = paste0("E", 1:6)
mat

df = data.frame(from = rep(rownames(mat), times = ncol(mat)),
    to = rep(colnames(mat), each = nrow(mat)),
    value = as.vector(mat),
    stringsAsFactors = FALSE)
df
```

The most simple usage is just calling `chordDiagram()` with 
`mat` (Figure \@ref(fig:chord-diagram-basic) left).

```{r chord_diagram_basic_simple, eval = FALSE}
chordDiagram(mat)
circos.clear()
```

or call with `df`:

```{r eval = FALSE}
chordDiagram(df)
circos.clear()
```

The default Chord Diagram consists of a track of labels, a track of grids (or you call it lines) with
axes, and links. Sectors which correspond to rows in the matrix locate at the
bottom half of the circle. The order of sectors is the order of
`union(rownames(mat), colnames(mat))` or `union(df[[1]], df[[2]])` if input is
a data frame. The order of sectors can be controlled by `order` argument
(Figure \@ref(fig:chord-diagram-basic) right). Of course, the length of
`order` vector should be same as the number of sectors.

```{r chord_diagram_basic_order, eval = FALSE}
chordDiagram(mat, order = c("S1", "E1", "E2", "S2", "E3", "E4", "S3", "E5", "E6"))
```

```{r chord-diagram-basic, echo = FALSE, fig.width = 8, fig.height = 4, fig.cap = "Basic usages of `chordDiagram()`."}
chunks <- knitr:::knit_code$get()
par(mfrow = c(1, 2))
eval(parse(text = chunks[["chord_diagram_basic_simple"]]))
eval(parse(text = chunks[["chord_diagram_basic_order"]]))
```

Under default settings, the grid colors which represent sectors are randomly
generated, and the link colors are same as grid colors which correspond to
rows but with 50% transparency.

## Adjust by `circos.par()`

Since Chord Diagram is implemented by basic circlize functions, like normal circular plot,
the layout can be customized by `circos.par()`.

The gaps between sectors can be set by `circos.par(gap.after = ...)` (Figure \@ref(fig:chord-diagram-par), left). 
It is useful when rows and columns belong to different categories. 
Please note since you change default graphical settings, you need to 
use `circos.clear()` in the end of the plot to reset it.

```{r chord_diagram_basic_gap_after, eval = FALSE}
circos.par(gap.after = c(rep(5, nrow(mat)-1), 15, rep(5, ncol(mat)-1), 15))
chordDiagram(mat)
circos.clear()
```

If the input is a data frame:

```{r, eval = FALSE}
circos.par(gap.after = c(rep(5, length(unique(df[[1]]))-1), 15, 
                         rep(5, length(unique(df[[2]]))-1), 15))
chordDiagram(df)
circos.clear()
```

Similar to a normal circular plot, the first sector (which is the first row in
the adjacency matrix or the first row in the adjacency list) starts from right
center of the circle and sectors are arranged clock wise. The start degree for
the first sector can be set by `circos.par(start.degree = ...)` and the
direction can be set by `circos.par(clock.wise = ...)` (Figure \@ref(fig:chord-diagram-par), right).

```{r chord_diagram_basic_start_degree, eval = FALSE}
circos.par(start.degree = 90, clock.wise = FALSE)
chordDiagram(mat)
circos.clear()
```

```{r chord-diagram-par, echo = FALSE, fig.width = 8, fig.height = 4, fig.cap = "Control `chordDiagram()` by `circos.par()`."}
chunks <- knitr:::knit_code$get()
par(mfrow = c(1, 2))
eval(parse(text = chunks[["chord_diagram_basic_gap_after"]]))
eval(parse(text = chunks[["chord_diagram_basic_start_degree"]]))
```

## Colors

### Set grid colors

Grids have different colors to represent different sectors. Generally, sectors
are divided into two groups. One contains sectors defined in rows of the
matrix or the first column in the data frame, and the other contains sectors
defined in columns of the matrix or the second column in the data frame. Thus,
links connect objects in the two groups. By default, link colors are same as
the color for the corresponding sectors in the first group.

Changing colors of grids may change the colors of links as well. Colors for
grids can be set through `grid.col` argument. Values of `grid.col` better be a
named vector of which names correspond to sector names. If it is has no name
index, the order of `grid.col` is assumed to have the same order as sectors 
(Figure \@ref(fig:chord-diagram-grid-color)).

```{r chord-diagram-grid-color, fig.width = 8, fig.height = 4, fig.cap = "Set grid colors in Chord diagram."}
grid.col = c(S1 = "red", S2 = "green", S3 = "blue",
    E1 = "grey", E2 = "grey", E3 = "grey", E4 = "grey", E5 = "grey", E6 = "grey")
chordDiagram(mat, grid.col = grid.col)
chordDiagram(t(mat), grid.col = grid.col)
```

### Set link colors {#chord-diagram-link-color}

Transparency of link colors can be set by `transparency` argument (Figure \ref{fig:chord-diagram-color} A). 
The value should between 0 to 1 in which 0 means no transparency and 1 means full transparency.
Default transparency is 0.5.

```{r chord_diagram_color_transparency, eval = FALSE, echo = 1}
chordDiagram(mat, grid.col = grid.col, transparency = 0)
text(-0.9, 0.9, "A", cex = 1.5)
```

For adjacecy matrix, colors for links can be customized by providing a matrix
of colors. In the following example, we use `rand_color()` to generate a
random color matrix. Note since `col_mat` already contains transparency,
`transparency` will be ignored if it is set (Figure \ref{fig:chord-diagram-color} B).

```{r chord_diagram_color_mat, eval = FALSE, echo = 1:3}
col_mat = rand_color(length(mat), transparency = 0.5)
dim(col_mat) = dim(mat)  # to make sure it is a matrix
chordDiagram(mat, grid.col = grid.col, col = col_mat)
text(-0.9, 0.9, "B", cex = 1.5)
```

While for ajacency list, colors for links can be customized as a vector.

```{r, eval = FALSE}
col = rand_color(nrow(df))
chordDiagram(df, grid.col = grid.col, col = col)
```

When the strength of the relation (e.g. correlations) represents as continuous values,
`col` can also be specified as a self-defined color mapping function. `chordDiagram()`
accepts a color mapping generated by `colorRamp2()` (Figure \ref{fig:chord-diagram-color} C). 

```{r chord_diagram_color_fun, eval = FALSE, echo = 1:2}
col_fun = colorRamp2(range(mat), c("#FFEEEE", "#FF0000"), transparency = 0.5)
chordDiagram(mat, grid.col = grid.col, col = col_fun)
text(-0.9, 0.9, "C", cex = 1.5)
```

The color mapping function also works for adjacency list, but it will be
applied to the third column in the data frame, so you need to make sure the
third column has the correct values.

```{r, eval = FALSE}
chordDiagram(df, grid.col = grid.col, col = col_fun)
```

When the input is a matrix, sometimes you don't need to generate the whole
color matrix. You can just provide colors which correspond to rows or columns
so that links from a same row/column will have the same color (Figure \ref{fig:chord-diagram-color} D, E). 
Here note values of colors can be set as numbers,
color names or hex code, same as in the base R graphics.

```{r chord_diagram_color_row_col, eval = FALSE, echo = c(1, 3)}
chordDiagram(mat, grid.col = grid.col, row.col = 1:3)
text(-0.9, 0.9, "D", cex = 1.5)
chordDiagram(mat, grid.col = grid.col, column.col = 1:6)
text(-0.9, 0.9, "E", cex = 1.5)
```

`row.col` and `column.col` is specifically designed for matrix. There is no
similar settings for ajacency list.

To emphasize again, transparency of links can be included in `col`, `row.col`
or `column.col`, if transparency is already set there, `transparency` argument
will be ignored.

In Section \@ref(highlight-links), we will introduce how to highlight subset
of links by only assigning colors to them.

```{r chord-diagram-color, echo = FALSE, fig.width = 8, fig.height = 8*3/2, fig.cap = "Control colors for Chord diagram."}
chunks <- knitr:::knit_code$get()
par(mfrow = c(2, 3))
eval(parse(text = chunks[["chord_diagram_color_transparency"]]))
eval(parse(text = chunks[["chord_diagram_color_mat"]]))
eval(parse(text = chunks[["chord_diagram_color_fun"]]))
eval(parse(text = chunks[["chord_diagram_color_row_col"]]))
```

## Link border {#chord-diagram-link-border}

`link.lwd`, `link.lty` and `link.border` control the line width, the line
style and the color of the link border. All these three parameters can be set
either a single scalar or a matrix if the input is adjacency matrix.

If it is set as a single scalar, it means all links share the same style (Figure \@ref(fig:chord-diagram-border) A).

```{r chord_diagram_style_scalar, eval = FALSE}
chordDiagram(mat, grid.col = grid.col, link.lwd = 2, link.lty = 2, link.border = "red")
```

If it is set as a matrix, it should have same dimension as `mat`
(Figure \@ref(fig:chord-diagram-border) B). 

```{r chord_diagram_style_fullmat, eval = FALSE}
lwd_mat = matrix(1, nrow = nrow(mat), ncol = ncol(mat))
lwd_mat[mat > 12] = 2
border_mat = matrix(NA, nrow = nrow(mat), ncol = ncol(mat))
border_mat[mat > 12] = "red"
chordDiagram(mat, grid.col = grid.col, link.lwd = lwd_mat, link.border = border_mat)
```

The matrix is not necessary to have same dimensions as in `mat`. It can also
be a sub matrix (Figure \@ref(fig:chord-diagram-border) C). For rows or
columns of which the corresponding values are not specified in the matrix,
default values are filled in. It must have row names and column names so that
the settings can be mapped to the correct links.

```{r chord_diagram_style_submatrix, eval = FALSE}
border_mat2 = matrix("black", nrow = 1, ncol = ncol(mat))
rownames(border_mat2) = rownames(mat)[2]
colnames(border_mat2) = colnames(mat)
chordDiagram(mat, grid.col = grid.col, link.lwd = 2, link.border = border_mat2)
```

To be more convenient, graphic parameters can be set as a three-column data
frame in which the first two columns correspond to row names and column names
in the matrix, and the third column corresponds to the graphic parameters
(Figure \@ref(fig:chord-diagram-border) D).

```{r chord_diagram_style_dataframe, eval = FALSE}
lty_df = data.frame(c("S1", "S2", "S3"), c("E5", "E6", "E4"), c(1, 2, 3))
lwd_df = data.frame(c("S1", "S2", "S3"), c("E5", "E6", "E4"), c(2, 2, 2))
border_df = data.frame(c("S1", "S2", "S3"), c("E5", "E6", "E4"), c(1, 1, 1))
chordDiagram(mat, grid.col = grid.col, link.lty = lty_df, link.lwd = lwd_df,
    link.border = border_df)
```

It is much easier if the input is a data frame, you only need to set graphic settings
as a vecotr.

```{r, eval = FALSE}
chordDiagram(df, grid.col = grid.col, link.lty = sample(1:3, nrow(df), replace = TRUE),
    link.lwd = runif(nrow(df))*2, link.border = sample(0:1, nrow(df), replace = TRUE))
```

```{r chord-diagram-border, echo = FALSE, fig.width = 8, rig.height = 8, fig.cap = "Control borders for Chord Diagram."}
chunks <- knitr:::knit_code$get()
par(mfrow = c(2, 2))
eval(parse(text = chunks[["chord_diagram_style_scalar"]]))
text(-0.9, 0.9, "A", cex = 1.5)
eval(parse(text = chunks[["chord_diagram_style_fullmat"]]))
text(-0.9, 0.9, "B", cex = 1.5)
eval(parse(text = chunks[["chord_diagram_style_submatrix"]]))
text(-0.9, 0.9, "C", cex = 1.5)
eval(parse(text = chunks[["chord_diagram_style_dataframe"]]))
text(-0.9, 0.9, "D", cex = 1.5)
```

## Highlight links

Sometimes we want to highlight some links to emphasize the importance of such relations.
Highlighting by different border styles are introduced in Section \@ref(chord-diagram-link-border) 
and here we focus on highlighting by colors.

THere are two ways to highlight links, one is to set different transparency to different links
and the other is to only draw links that needs to be highlighted. Based on this rule and 
ways to assign colors to links (introduced in Section \@ref(chord-diagram-link-color)), We can
highlight links which come from a same sector by assigning colors with different transparency
by `row.col` argument (Figure \@ref(fig:chord-diagram-highlight) A).

```{r chord_diagram_highlight_row, eval = FALSE}
chordDiagram(mat, grid.col = grid.col, row.col = c("#FF000080", "#00FF0010", "#0000FF10"))
```

We can also highlight links with values larger than a cutoff (Figure \@ref(fig:chord-diagram-highlight) B). 
There are at least three ways to do it. First, construct a color matrix and set links
with small values to full transparency.

Since link colors can be specified as a matrix, we can set the transparency of those links to a high value
or even set to full transparency. In following example, links with value less then 12 is set to `#00000000`.

```{r chord_diagram_highlight_mat, eval = FALSE}
col_mat[mat < 12] = "#00000000"
chordDiagram(mat, grid.col = grid.col, col = col_mat) 
```

Following code demonstrates using a color mapping function to map values to different transparency.
Note this is also workable for adjacency list.

```{r chord_diagram_highlight_fun, eval = FALSE}
col_fun = function(x) ifelse(x < 12, "#00000000", "#FF000080")
chordDiagram(mat, grid.col = grid.col, col = col_fun)
```

For both color matrix and color mapping function, actually all links are all
drawn and the reason why you cannot see some of them is they are assigned with
full transparency. If a three-column data frame is used to assign colors to
links of interest, links which are not defined in `col_df` are not drawn (Figure \@ref(fig:chord-diagram-highlight) C).

```{r chord_diagram_highlight_df, eval = FALSE}
col_df = data.frame(c("S1", "S2", "S3"), c("E5", "E6", "E4"), 
    c("#FF000080", "#00FF0080", "#0000FF80"))
chordDiagram(mat, grid.col = grid.col, col = col_df)
```

Highlighting links is relatively simple for adjacency list that you only need
to construct a vector of colors according to what links you want to highlight.

```{r, eval = FALSE}
col = rand_color(nrow(df))
col[df[[3]] < 10] = "#00000000"
chordDiagram(df, grid.col = grid.col, col = col)
```

The `link.visible` argument is recently introduced to **circlize** package
which may provide a simple to control the visibility of links. The value can
be set as an logical matrix for adjacency matrix or a logical vector for
adjacency list (Figure \@ref(fig:chord-diagram-highlight) D).

```{r chord_diagram_visible, eval = FALSE}
col = rand_color(nrow(df))
chordDiagram(df, grid.col = grid.col, link.visible = df[[3]] >= 10)
```

```{r chord-diagram-highlight, echo = FALSE, fig.width = 8, fig.height = 8, fig.cap = "Highlight links by colors."}
chunks <- knitr:::knit_code$get()
par(mfrow = c(2, 2))
eval(parse(text = chunks[["chord_diagram_highlight_row"]]))
text(-0.9, 0.9, "A", cex = 1.5)
eval(parse(text = chunks[["chord_diagram_highlight_mat"]]))
text(-0.9, 0.9, "B", cex = 1.5)
eval(parse(text = chunks[["chord_diagram_highlight_df"]]))
text(-0.9, 0.9, "C", cex = 1.5)
eval(parse(text = chunks[["chord_diagram_visible"]]))
text(-0.9, 0.9, "D", cex = 1.5)
```

## Orders of links

### Orders on positions on sectors

Orders of links on every sector are adjusted automatically to make them look
nice. But sometimes sorting links according to the width on the sector is
useful for detecting potential features. `link.sort` and `link.decreasing` can
be set to control the order of positioning links on sectors 
(Figure \@ref(fig:chord-diagram-link-order1)).

```{r chord-diagram-link-order1, echo = c(2, 4), fig.width = 8, fig.height = 4, fig.cap = "Order of positioning links on sectors."}
par(mfrow = c(1, 2))
chordDiagram(mat, grid.col = grid.col, link.border = 1, link.sort = TRUE, link.decreasing = TRUE)
text(-0.9, 0.9, "A", cex = 1.5)
chordDiagram(mat, grid.col = grid.col, link.border = 1, link.sort = TRUE, link.decreasing = FALSE)
text(-0.9, 0.9, "B", cex = 1.5)
```

### Order of adding links

The default order of adding links to the plot is based on their order in the
matrix or in the data frame. Normally, transparency should be set to the link
colors so that they will not overlap to each other. In most cases, this looks
fine, but sometimes, it improves the visualization to put wide links more
forward and to put small links more backward in the plot. This can be set by
`link.rank` argument which defines the order of adding links. Larger value
means the corresponding link is added later (Figure \@ref(fig:chord-diagram-link-order2)).

```{r chord-diagram-link-order2, echo = c(2, 4), fig.width = 8, fig.height = 4, fig.cap = "Order of adding links."}
chordDiagram(df, grid.col = grid.col, transparency = 1)
text(-0.9, 0.9, "A", cex = 1.5)
chordDiagram(df, grid.col = grid.col, transparency = 1, link.rank = rank(df[[3]]))
text(-0.9, 0.9, "B", cex = 1.5)
```

## Self-links

How to set self links dependends on whether the information needs to be duplicated. 
The `self.link` argument can be set to `1` or `2` for the two different scenarios. 
Check the difference in Figure \@ref(fig:chord-diagram-self-link).

```{r chord-diagram-self-link, fig.width = 8, fig.height = 4, fig.cap = "Self-links in Chord diagram."}
df2 = data.frame(start = c("a", "b", "c", "a"), end = c("a", "a", "b", "c"))
chordDiagram(df2, grid.col = 1:3, self.link = 1)
chordDiagram(df2, grid.col = 1:3, self.link = 2)
```

## Symmetric matrix

When the matrix is symmetric, by setting `symmetric = TRUE`, only lower
triangular matrix without the diagonal will be used (Figure \@ref(fig:chord-diagram-symmetric)).

```{r chord-diagram-symmetric, echo = 2:7, fig.width = 8, fig.height = 4, fig.cap = "Symmetric matrix for Chord diagram."}
par(mfrow = c(1, 2))
mat3 = matrix(rnorm(25), 5)
colnames(mat3) = letters[1:5]
cor_mat = cor(mat3)
col_fun = colorRamp2(c(-1, 0, 1), c("green", "white", "red"))
chordDiagram(cor_mat, grid.col = 1:5, symmetric = TRUE, col = col_fun)
chordDiagram(cor_mat, grid.col = 1:5, col = col_fun)
```

## Directional relations

In some cases, when the input is a matrix, rows and columns represent
directions, or when the input is a data frame, the first column and second
column represent directions. Argument `directional` is used to illustrate such
direction on the plot. `directional` with value `1` means the direction is
from rows to columns (or from the first column to the second column for the
adjacency list) while `-1` means the direction is from columns to rows (or from
the second column to the first column for the adjacency list). A value of `2` means
bi-directional.

By default, the two ends of links have unequal height (Figure
\@ref(fig:chord-diagram-directional) A, B, C) to represent the directions.
The position of starting end of the link is shorter than the other end to give
users the feeling that the link is moving out. If this is not what your
correct feeling, you can set `diffHeight` to a negative value.


```{r chord_diagram_directional_simple, eval = FALSE, echo = c(1, 3, 5)}
chordDiagram(mat, grid.col = grid.col, directional = 1)
text(-0.9, 0.9, "A", cex = 1.5)
chordDiagram(mat, grid.col = grid.col, directional = 1, diffHeight = 0.08)
text(-0.9, 0.9, "B", cex = 1.5)
chordDiagram(mat, grid.col = grid.col, directional = -1)
text(-0.9, 0.9, "C", cex = 1.5)
```

Row names and column names in `mat` can also overlap. In this case, showing
direction of the link is important to distinguish them (Figure \@ref(fig
:chord-diagram-directional) D).

```{r}
mat2 = matrix(sample(100, 35), nrow = 5)
rownames(mat2) = letters[1:5]
colnames(mat2) = letters[1:7]
mat2
```

```{r chord_diagram_directional_overlap, eval = FALSE}
chordDiagram(mat2, grid.col = 1:7, directional = 1, row.col = 1:5)
```

If you don't need self-link for which two ends of a link are in a same sector, 
just set corresponding values to 0 in the matrix (Figure \@ref(fig:chord-diagram-directional) E).

```{r}
mat3 = mat2
for(cn in intersect(rownames(mat3), colnames(mat3))) {
    mat3[cn, cn] = 0
}
mat3
```

```{r chord_diagram_directional_non_selfloop, eval = FALSE}
chordDiagram(mat3, grid.col = 1:7, directional = 1, row.col = 1:5)
```

Links can have arrows to represent directions (Figure \@ref(fig:chord-diagram-directional) F). 
When `direction.type` is set to `arrows`, Arrows are added at
the center of links. Similar as other graphics parameters for links, the
parameters for drawing arrows such as arrow color and line type can either be
a scalar, a matrix, or a three-column data frame.

If `link.arr.col` is set as a data frame, only links specified in the data frame
will have arrows. Pleast note this is the only way to draw arrows to subset of links.

```{r chord_diagram_directional_arrow, eval = FALSE}
arr.col = data.frame(c("S1", "S2", "S3"), c("E5", "E6", "E4"), 
    c("black", "black", "black"))
chordDiagram(mat, grid.col = grid.col, directional = 1, direction.type = "arrows",
    link.arr.col = arr.col, link.arr.length = 0.2)
```

If combining both `arrows` and `diffHeight`, it will give you better visualization 
(Figure \@ref(fig:chord-diagram-directional) G).

```{r chord_diagram_directional_arrow2, eval = FALSE}
arr.col = data.frame(c("S1", "S2", "S3"), c("E5", "E6", "E4"), 
    c("black", "black", "black"))
chordDiagram(mat, grid.col = grid.col, directional = 1, 
    direction.type = c("diffHeight", "arrows"),
    link.arr.col = arr.col, link.arr.length = 0.2)
```

There is another arrow type: `big.arrow` which is efficient to visualize
arrows when there are too many links (Figure \@ref(fig:chord-diagram-directional) H)

```{r chord_diagram_directional_arrow3, eval = FALSE}
matx = matrix(rnorm(64), 8)
chordDiagram(matx, directional = 1, direction.type = c("diffHeight", "arrows"),
    link.arr.type = "big.arrow")
```

If `diffHeight` is set to a negative value, the start ends are longer than
the other ends (Figure \@ref(fig:chord-diagram-directional) I)

```{r chord_diagram_directional_arrow4, eval = FALSE}
chordDiagram(matx, directional = 1, direction.type = c("diffHeight", "arrows"),
    link.arr.type = "big.arrow", diffHeight = -0.04)
```

It is almost the same to visualize directional Chord diagram form a adjacency list.

```{r, eval = FALSE}
# code is not run when building the vignette
chordDiagram(df, directional = 1)
```

```{r chord-diagram-directional, fig.width = 8, fig.height = 8, fig.cap = "Visualization of directional matrix."}
par(mfrow = c(3, 3))
eval(parse(text = chunks[["chord_diagram_directional_simple"]]))
eval(parse(text = chunks[["chord_diagram_directional_overlap"]]))
eval(parse(text = chunks[["chord_diagram_directional_non_selfloop"]]))
eval(parse(text = chunks[["chord_diagram_directional_arrow"]]))
eval(parse(text = chunks[["chord_diagram_directional_arrow2"]]))
eval(parse(text = chunks[["chord_diagram_directional_arrow3"]]))
eval(parse(text = chunks[["chord_diagram_directional_arrow4"]]))
```


