```{r, echo = FALSE}
library(circlize)
```

# A complex example of Chord diagram

In this chapter, we demonstrate how to make a complex Chord diagram and how to
customize additional tracks by visualizing chromatin state transitions as well
as methylation changes. A chromatin state transition matrix shows how much a
chromatin state in the genome has been changed from e.g. one group of samples
to the other. The genomic regions for which the chromatin states change also
have variable methylation patterns which may show interesting correspondance
to chromatin states change.

The data used in this post is processed from [Roadmap
dataset](http://www.roadmapepigenomics.org/). The chromatin states [are learned
from five core chromatin marks](http://egg2.wustl.edu/roadmap/web_portal/chr_s
tate_learning.html#core_15state). Roadmap samples are classified into two
groups based on expression profile. In each group, a chromatin state is
assigned to the corresponding genomic bin if it is recurrent in at least half
of the samples in each group.

The processed data can be set from http://jokergoo.github.io/data/chromatin_transition.RData.

```{r}
load("data/chromatin_transition.RData")
```

In the RData file, there are three matrix: `mat`, `meth_mat_1` and
`meth_mat_2` which are:

- `mat`: chromatin state transition matrix. Rows correspond to states in group
  1 and columns correspond to group 2. The value in the matrix are total base
  pairs that transite from one state to the other. E.g. `mat["TssA",
  "TxFlnk"]` is the total base pairs that have "TssA" state in samples in
  group 1 and transites to "TxFlnk" state in samples in group 2. On the
  diagonal are the regions where the states have not been changed in the two
  groups.
- `meth_mat_1`: mean methylation for each set of regions in group 1.
  `meth_mat_1["TssA", "TxFlnk"]` is the mean methylation for the regions in
  **group 1** that have "TssA" state in group 1 and "TxFlnk" state in group 2.
- `meth_mat_2`: mean methylation for each set of regions in group 2.
  `meth_mat_2["TssA", "TxFlnk"]` is the mean methylation for the regions in
  **group 2** that have "TssA" state in group 1 and "TxFlnk" state in group 2.

```{r}
mat[1:4, 1:4]
meth_mat_1[1:4, 1:4]
```

Normally, in majority in the genome, chromatin states of regions are not
changed in the two groups, thus, we should only look at the regions in which
the states are changed.

```{r}
# proportion of the unchanges states in the genome
sum(diag(mat))/sum(mat)
# remove the unchanged states
diag(mat) = 0
```

When making the plot, actually rows and columns are different (because one is
from group 1 and the other is from group 2), thus we give them different names
and the original names are stored in `all_states`.

```{r}
all_states = rownames(mat)
n_states = nrow(mat)

rownames(mat) = paste0("R_", seq_len(n_states))
colnames(mat) = paste0("C_", seq_len(n_states))

dimnames(meth_mat_1) = dimnames(mat)
dimnames(meth_mat_2) = dimnames(mat)
```

Next we set the colors. `colmat` is the color of the links and the colors
are represent as hexadecimal code. Links have more transparent (`A0`) if they
contain few transitions (< 70th percentile) because we don't want it
to disturb the visualization of the major transitions.

```{r}
state_col = c("TssA" = "#E41A1C",
	          "TssAFlnk" = "#E41A1C",
	          "TxFlnk" = "#E41A1C",
	          "Tx" = "#E41A1C",
	          "TxWk" = "#E41A1C",
	          "EnhG" = "#E41A1C",
	          "Enh" = "#E41A1C",
	          "ZNF/Rpts" = "#E41A1C",
	          "Het" = "#377EB8",
	          "TssBiv" = "#377EB8",
	          "BivFlnk" = "#377EB8",
	          "EnhBiv" = "#377EB8",
	          "ReprPC" = "#377EB8",
	          "ReprPCWk" = "#377EB8",
	          "Quies" = "black")

# one for rows and one for columns
state_col2 = c(state_col, state_col)
names(state_col2) = c(rownames(mat), colnames(mat))

colmat = rep(state_col2[rownames(mat)], n_states)
colmat = rgb(t(col2rgb(colmat)), maxColorValue = 255)

qati = quantile(mat, 0.7)
colmat[mat > qati] = paste0(colmat[mat > qati], "A0")
colmat[mat <= qati] = paste0(colmat[mat <= qati], "20")
dim(colmat) = dim(mat)
```

Now we can use `chordDiagram()` function to make the plot. Here we set one
pre-allocated track in which the methylation information will be added later.
Also we only set `annotationTrack` to `grid` and the axes and sector labels
will be customized in later code.

`chordDiagram()` returns a data frame which contains coordinates for all links.

```{r first, fig.width = 6, fig.height = 6}
circos.par(start.degree = -5, gap.after = c(rep(1, n_states-1), 10, rep(1, n_states-1), 10),
	cell.padding = c(0, 0, 0, 0), points.overflow.warning = FALSE)

cdm_res = chordDiagram(mat, col = colmat, grid.col = state_col2,
	directional = TRUE, annotationTrack = "grid", preAllocateTracks = list(track.height = 0.1))
head(cdm_res)
```

```{r, echo = FALSE}
circos.clear()
chunks <- knitr:::knit_code$get()
```

Now the axes are added in the second track, also, the index of states are
added at the center of the grids in the second track, if the degree for a
sector is larger than 3 degrees. Note since there is already one pre-allocated
track, the circular rectangles are in the second track (`track.index = 2`).

```{r second, eval = FALSE}
circos.track(track.index = 2, panel.fun = function(x, y) {
	if(abs(CELL_META$cell.start.degree - CELL_META$cell.end.degree) > 3) {
		sn = CELL_META$sector.index
		i_state = as.numeric(gsub("(C|R)_", "", sn))
		circos.text(CELL_META$xcenter, CELL_META$ycenter, i_state, col = "white", 
			font = 2, cex = 0.7, adj = c(0.5, 0.5), niceFacing = TRUE)
		xlim = CELL_META$xlim
		breaks = seq(0, xlim[2], by = 4e5)
		circos.axis(major.at = breaks, labels = paste0(breaks/1000, "KB"), labels.cex = 0.5)
	}
}, bg.border = NA)
```

```{r, results = "hide", fig.width = 6, fig.height = 6, echo = FALSE}
eval(parse(text = chunks[["first"]]))
eval(parse(text = chunks[["second"]]))
circos.clear()
```

On the top half, it is easy to see the proportion of different transitions in
group 1 that come to every state in group 2. However, it is not
straightforward for the states in the bottom half to see the proportion of
different states in group 2 they transite to. This can be solved by adding
small circular rectangles. In following example, the newly added circular
rectangles in the bottom half shows e.g. how much the state 15 in group 1 has
been transited to different states in group 2.

```{r third, eval = FALSE}
for(i in seq_len(nrow(cdm_res))) {
	if(cdm_res$value[i] > 0) {
		circos.rect(cdm_res[i, "x1"], -uy(1, "mm"), 
			cdm_res[i, "x1"] - abs(cdm_res[i, "value"]), -uy(2, "mm"), 
			col = state_col2[cdm_res$cn[i]], border = state_col2[cdm_res$cn[i]],
			sector.index = cdm_res$rn[i], track.index = 2)
	}
}
```

```{r, results = "hide", fig.width = 6, fig.height = 6, echo = FALSE}
eval(parse(text = chunks[["first"]]))
eval(parse(text = chunks[["second"]]))
eval(parse(text = chunks[["third"]]))
circos.clear()
```

Methylation in each category is put on the most outside of the circle. On this track, we will
put two paralle rectangles which are mean methylation and methylation difference between group 1
and group 2. Basically, on the bottom, we show `meth_mat_2 - meth_mat_1` and on the top we show
`meth_mat_1 - meth_mat_2`.

```{r last, eval = FALSE}
abs_max = quantile(abs(c(meth_mat_1, meth_mat_2) - 0.5), 0.95, na.rm = TRUE)
col_fun = colorRamp2(c(0.5 - abs_max, 0.5, 0.5 + abs_max), c("blue", "white", "red"))
col_fun2 = colorRamp2(c(-abs_max, 0, abs_max), c("green", "white", "orange"))

ylim = get.cell.meta.data("ylim", sector.index = rownames(mat)[1], track.index = 1)
y1 = ylim[1] + (ylim[2] - ylim[1])*0.4
y2 = ylim[2]
for(i in seq_len(nrow(cdm_res))) {
	if(cdm_res$value[i] > 0) {
		circos.rect(cdm_res[i, "x1"], y1, cdm_res[i, "x1"] - abs(cdm_res[i, "value"]), y1 + (y2-y1)*0.45, 
			col = col_fun(meth_mat_1[cdm_res$rn[i], cdm_res$cn[i]]), 
			border = col_fun(meth_mat_1[cdm_res$rn[i], cdm_res$cn[i]]),
			sector.index = cdm_res$rn[i], track.index = 1)

		circos.rect(cdm_res[i, "x1"], y1 + (y2-y1)*0.55, cdm_res[i, "x1"] - abs(cdm_res[i, "value"]), y2, 
			col = col_fun2(meth_mat_2[cdm_res$rn[i], cdm_res$cn[i]] - meth_mat_1[cdm_res$rn[i], cdm_res$cn[i]]), 
			border = col_fun2(meth_mat_2[cdm_res$rn[i], cdm_res$cn[i]] - meth_mat_1[cdm_res$rn[i], cdm_res$cn[i]]),
			sector.index = cdm_res$rn[i], track.index = 1)

		circos.rect(cdm_res[i, "x2"], y1, cdm_res[i, "x2"] - abs(cdm_res[i, "value"]), y1 + (y2-y1)*0.45, 
			col = col_fun(meth_mat_2[cdm_res$rn[i], cdm_res$cn[i]]), 
			border = col_fun(meth_mat_2[cdm_res$rn[i], cdm_res$cn[i]]),
			sector.index = cdm_res$cn[i], track.index = 1)

		circos.rect(cdm_res[i, "x2"], y1 + (y2-y1)*0.55, cdm_res[i, "x2"] - abs(cdm_res[i, "value"]), y2, 
			col = col_fun2(meth_mat_1[cdm_res$rn[i], cdm_res$cn[i]] - meth_mat_2[cdm_res$rn[i], cdm_res$cn[i]]), 
			border = col_fun2(meth_mat_1[cdm_res$rn[i], cdm_res$cn[i]] - meth_mat_2[cdm_res$rn[i], cdm_res$cn[i]]),
			sector.index = cdm_res$cn[i], track.index = 1)
	}
}
circos.clear()
```


```{r, results = "hide", fig.width = 6, fig.height = 6, echo = FALSE}
eval(parse(text = chunks[["first"]]))
eval(parse(text = chunks[["second"]]))
eval(parse(text = chunks[["third"]]))
eval(parse(text = chunks[["last"]]))
circos.clear()
```

Legends can be added according to instructions discussed in Section \@ref(legends).

```{r, echo = FALSE, fig.height = 6, fig.width = 6 + 1.165229 + 0.08}
library(grid)
library(gridBase)
plot.new()
circle_size = unit(1, "snpc") # snpc unit gives you a square region

pushViewport(viewport(x = 0, y = 0.5, width = circle_size, height = circle_size,
    just = c("left", "center")))
par(omi = gridOMI(), new = TRUE)
eval(parse(text = chunks[["first"]]))
eval(parse(text = chunks[["second"]]))
eval(parse(text = chunks[["third"]]))
eval(parse(text = chunks[["last"]]))
circos.clear()
upViewport()

library(ComplexHeatmap)
lgd_list_vertical = packLegend(
	Legend(at = 1:n_states, labels = names(state_col), pch = as.character(1:n_states),, 
		legend_gp = gpar(col = "white", fontsize = 8), background = state_col, type = "points", 
		title = "Chromatin States"),
	Legend(at = c(0.1, 0.3, 0.5, 0.7, 0.9), col_fun = col_fun, title = "Mean Methylation"),
	Legend(at = c(-0.4, -0.2, 0, 0.2, 0.4), col_fun = col_fun2, title = "Mean Difference"))
pushViewport(viewport(x = circle_size, y = 0.5, width = grobWidth(lgd_list_vertical), 
    height = grobHeight(lgd_list_vertical), just = c("left", "center")))
grid.draw(lgd_list_vertical)
upViewport()
```
