```{r, echo = FALSE}
library(circlize)
```

# A complex example of Chord diagram

Visualize chromatin state transitions
=========================================

A chromatin state transition matrix shows how much the chromatin state in the genome
has been changed from e.g. one sample to another. In our [recent paper](http://msb.embopress.org/content/12/3/861) 
we demonstrated how chromatin states change between smoking people and non-smoking people 
by means of [Chord Diagram](http://d3dwu2jylmmhzr.cloudfront.net/content/msb/12/3/861/F8.large.jpg).
In this post, I will demonstrate how to make such plot by the **circlize** package 
and enhance it by adding methylation information.

The data demonstrated in this post is processed from [Roadmap data](http://www.roadmapepigenomics.org/). The
chromatin states [are learned from five core chromatin marks](http://egg2.wustl.edu/roadmap/web_portal/chr_state_learning.html#core_15state). Roadmap samples are separated into two groups based on expression profile. In each group,
a chromatin state is assigned to the corresponding genomic bin if it is recurrent in at least half of the samples.

The processed data is stored as [chromatin_transition.RData](../../data/chromatin_transition.RData).

```{r}
library(circlize)
load("../data/chromatin_transition.RData")
```

There are three matrix: `mat`, `meth_mat_1` and `meth_mat_2` which are:

- `mat`: chromatin state transition matrix. Rows correspond to states in group 1 and columns
   correspond to group 2. The value in the matrix are total base pairs that transite from one group to the other.
   E.g. there are in total `r as.integer(mat[1, 2])` bp which are in "`r rownames(mat)[1]`" state in group 1 and they change 
   to "`r colnames(mat)[2]`" state in group 2. On the digonal are the unchanged states.
- `meth_mat_1`: mean methylation in group 1 in each category.
- `meth_mat_2`: mean methylation in group 2 in each category.

```{r}
mat[1:4, 1:4]
meth_mat_1[1:4, 1:4]
```

Normally, majority in the genome are unchanged states, thus, we should only look at the regions
in which their states are changed.

```{r}
# proportion of the unchanges states in the genome
sum(diag(mat))/sum(mat)
# remove the unchanged states
diag(mat) = 0
```

When making the plot, actually rows and columns are different (because one is from group 1 and the other is from group 2),
thus we give them different names and the original names are stored in `all_states`.

```{r}
all_states = rownames(mat)
n_states = nrow(mat)

rownames(mat) = paste0("R_", seq_len(n_states))
colnames(mat) = paste0("C_", seq_len(n_states))

dimnames(meth_mat_1) = dimnames(mat)
dimnames(meth_mat_2) = dimnames(mat)
```

Next we set the colors. `colmat` is the color of the links and the colors
are represent as hex code. Links have more transparent (`A0`) if they
contain few transitions (< 70th percentile) because we don't want it
to disturb the visualization of the major transitions.

```{r}
state_col = c("TssA" = "#E41A1C",
	          "TssAFlnk" = "#E41A1C",
	          "TxFlnk" = "#E41A1C",
	          "Tx" = "#E41A1C",
	          "TxWk" = "#E41A1C",
	          "EnhG" = "#E41A1C",
	          "Enh" = "#E41A1C",
	          "ZNF/Rpts" = "#E41A1C",
	          "Het" = "#377EB8",
	          "TssBiv" = "#377EB8",
	          "BivFlnk" = "#377EB8",
	          "EnhBiv" = "#377EB8",
	          "ReprPC" = "#377EB8",
	          "ReprPCWk" = "#377EB8",
	          "Quies" = "black")

# one for rows and one for columns
state_col2 = c(state_col, state_col)
names(state_col2) = c(rownames(mat), colnames(mat))

colmat = rep(state_col2[rownames(mat)], n_states)
colmat = rgb(t(col2rgb(colmat)), maxColorValue = 255)

qati = quantile(mat, 0.7)
colmat[mat > qati] = paste0(colmat[mat > qati], "A0")
colmat[mat <= qati] = paste0(colmat[mat <= qati], "20")
dim(colmat) = dim(mat)
```

Now we can use `chordDiagram()` function to make the plot. Here we set one pre-allocated
track in which the methylation information will be put.

`chordDiagram()` returns a data frame which contains coordinates for all links which will
be used later.

`de` is the degree for the "gap" between group 1 and group 2.

```{r first, eval = FALSE}
de = 360 - (360 - 20 - 30) - 30
circos.par(start.degree = -de/4, gap.degree = c(rep(1, n_states-1), de/2, rep(1, n_states-1), de/2))

cdm_res = chordDiagram(mat, col = colmat, grid.col = state_col2,
	directional = TRUE, annotationTrack = "grid", preAllocateTracks = list(track.height = 0.1))
```

```{r, results = "hide", fig.width = 8, fig.height = 8, echo = FALSE}
chunks <- knitr:::knit_code$get()
eval(parse(text = chunks[["first"]]))
circos.clear()
```

If the degree for a sector is larger than 3 degrees, the index for the state and axis is added.
Note since there is already one pre-allocated track, the circular rectangles are in the second track (`track.index = 2`).

```{r second, eval = FALSE}
for(sn in get.all.sector.index()) {
	if(abs(get.cell.meta.data("cell.start.degree", sector.index = sn) - 
	       get.cell.meta.data("cell.end.degree", sector.index = sn)) > 3) {
		xcenter = get.cell.meta.data("xcenter", sector.index = sn, track.index = 2)
		ycenter = get.cell.meta.data("ycenter", sector.index = sn, track.index = 2)
		i_state = as.numeric(gsub("(C|R)_", "", sn))
		circos.text(xcenter, ycenter, i_state, col = "white", font = 2, cex = 0.7, 
			sector.index = sn, track.index = 2, adj = c(0.5, 0.5), niceFacing = TRUE)
		circos.axis(sector.index = sn, track.index = 2, major.tick.percentage = 0.2, 
			labels.away.percentage = 0.2, labels.cex = 0.5)
	}
}
```

```{r, results = "hide", fig.width = 8, fig.height = 8, echo = FALSE}
eval(parse(text = chunks[["first"]]))
eval(parse(text = chunks[["second"]]))
circos.clear()
```

On the top half, it is easy to see the proportion of different transitions in group 1 that come to every state in group 2.
However, it is not straightforward for the states in the bottom half to see the proportion of different
states in group 2 they transite to. This can be solved by adding small circular rectangles. In following example, 
the newly added circular rectangles in the bottom half shows e.g. how much the state 15 in group 1 has been transited
to different states in group 2.

```{r third, eval = FALSE}
for(i in seq_len(nrow(cdm_res))) {
	if(cdm_res$value[i] > 0) {
		circos.rect(cdm_res[i, "x1"], -0.5, cdm_res[i, "x1"] - abs(cdm_res[i, "value"]), -0.7, 
			col = state_col2[cdm_res$cn[i]], border = state_col2[cdm_res$cn[i]],
			sector.index = cdm_res$rn[i], track.index = 2)
	}
}
```

```{r, results = "hide", fig.width = 8, fig.height = 8, echo = FALSE}
eval(parse(text = chunks[["first"]]))
eval(parse(text = chunks[["second"]]))
eval(parse(text = chunks[["third"]]))
circos.clear()
```

Methylation in each category is put on the most outside of the circle. On this track, we will
put two paralle rectangles which are mean methylation and methylation difference between group 1
and group 2. Basically, on the bottom, we show `meth_mat_2 - meth_mat_1` and on the top we show
`meth_mat_1 - meth_mat_2`.

```{r last, eval = FALSE}
abs_max = quantile(abs(c(meth_mat_1, meth_mat_2) - 0.5), 0.95, na.rm = TRUE)
col_fun = colorRamp2(c(0.5 - abs_max, 0.5, 0.5 + abs_max), c("blue", "white", "red"))
col_fun2 = colorRamp2(c(-abs_max, 0, abs_max), c("green", "white", "orange"))

ylim = get.cell.meta.data("ylim", sector.index = rownames(mat)[1], track.index = 1)
y1 = ylim[1] + (ylim[2] - ylim[1])*0.4
y2 = ylim[2]
for(i in seq_len(nrow(cdm_res))) {
	if(cdm_res$value[i] > 0) {
		circos.rect(cdm_res[i, "x1"], y1, cdm_res[i, "x1"] - abs(cdm_res[i, "value"]), y1 + (y2-y1)*0.45, 
			col = col_fun(meth_mat_1[cdm_res$rn[i], cdm_res$cn[i]]), 
			border = col_fun(meth_mat_1[cdm_res$rn[i], cdm_res$cn[i]]),
			sector.index = cdm_res$rn[i], track.index = 1)

		circos.rect(cdm_res[i, "x1"], y1 + (y2-y1)*0.55, cdm_res[i, "x1"] - abs(cdm_res[i, "value"]), y2, 
			col = col_fun2(meth_mat_2[cdm_res$rn[i], cdm_res$cn[i]] - meth_mat_1[cdm_res$rn[i], cdm_res$cn[i]]), 
			border = col_fun2(meth_mat_2[cdm_res$rn[i], cdm_res$cn[i]] - meth_mat_1[cdm_res$rn[i], cdm_res$cn[i]]),
			sector.index = cdm_res$rn[i], track.index = 1)

		circos.rect(cdm_res[i, "x2"], y1, cdm_res[i, "x2"] - abs(cdm_res[i, "value"]), y1 + (y2-y1)*0.45, 
			col = col_fun(meth_mat_2[cdm_res$rn[i], cdm_res$cn[i]]), 
			border = col_fun(meth_mat_2[cdm_res$rn[i], cdm_res$cn[i]]),
			sector.index = cdm_res$cn[i], track.index = 1)

		circos.rect(cdm_res[i, "x2"], y1 + (y2-y1)*0.55, cdm_res[i, "x2"] - abs(cdm_res[i, "value"]), y2, 
			col = col_fun2(meth_mat_1[cdm_res$rn[i], cdm_res$cn[i]] - meth_mat_2[cdm_res$rn[i], cdm_res$cn[i]]), 
			border = col_fun2(meth_mat_1[cdm_res$rn[i], cdm_res$cn[i]] - meth_mat_2[cdm_res$rn[i], cdm_res$cn[i]]),
			sector.index = cdm_res$cn[i], track.index = 1)
	}
}

circos.clear()
```


```{r, results = "hide", fig.width = 8, fig.height = 8, echo = FALSE}
eval(parse(text = chunks[["first"]]))
eval(parse(text = chunks[["second"]]))
eval(parse(text = chunks[["third"]]))
eval(parse(text = chunks[["last"]]))
circos.clear()
```

A complete plot with legends looks like:

```{r, results = "hide", fig.width = 8, fig.height = 8, echo = FALSE}
eval(parse(text = chunks[["first"]]))
eval(parse(text = chunks[["second"]]))
eval(parse(text = chunks[["third"]]))
eval(parse(text = chunks[["last"]]))
circos.clear()


add_chord_diagram_legend = function(position = c("bottomleft", "bottomright", "topleft", "topright"), 
	index = seq_along(labels), labels, col) {
	
	position = match.arg(position)[1]
	if(length(index) == 0) {
		return(NULL)
	}

	coor = par("usr")
	n = length(labels)
	text_height = strheight("a")
	labels_max_width = max(strwidth(labels))
	legend_width = text_height*(1+0.5) + labels_max_width
	if(position == "bottomleft") {
		x1 = rep(coor[1], n)
		x2 = x1 + text_height
		y1 = coor[3] + (rev(seq_len(n))-1)*1.5*text_height
		y2 = y1 + text_height
		rect(x1, y1, x2, y2, col = col, border = col)
		text((x1+x2)/2, (y1+y2)/2, index, cex = 0.6, font = 2, col = "white")
		text(x2 + 0.5*text_height, (y1+y2)/2, labels, adj = c(0, 0.5), cex = 0.8)
	} else if(position == "bottomright") {
		x1 = rep(coor[2] - labels_max_width, n)
		x2 = x1 + text_height
		y1 = coor[3] + (rev(seq_len(n))-1)*1.5*text_height
		y2 = y1 + text_height
		rect(x1, y1, x2, y2, col = col, border = col)
		text((x1+x2)/2, (y1+y2)/2, index, cex = 0.6, font = 2, col = "white")
		text(x2 + 0.5*text_height, (y1+y2)/2, labels, adj = c(0, 0.5), cex = 0.8)
	} else if(position == "topleft") {
		x1 = rep(coor[1], n)
		x2 = x1 + text_height
		y1 = coor[4] - (seq_len(n)-1)*1.5*text_height
		y2 = y1 - text_height
		rect(x1, y1, x2, y2, col = col, border = col)
		text((x1+x2)/2, (y1+y2)/2, index, cex = 0.6, font = 2, col = "white")
		text(x2 + 0.5*text_height, (y1+y2)/2, labels, adj = c(0, 0.5), cex = 0.8)
	} else if(position == "topright") {
		x1 = rep(coor[2] - labels_max_width, n)
		x2 = x1 + text_height
		y1 = coor[4] - (seq_len(n)-1)*1.5*text_height
		y2 = y1 - text_height
		rect(x1, y1, x2, y2, col = col, border = col)
		text((x1+x2)/2, (y1+y2)/2, index, cex = 0.6, font = 2, col = "white")
		text(x2 + 0.5*text_height, (y1+y2)/2, labels, adj = c(0, 0.5), cex = 0.8)
	}
}

ib = ceiling(n_states/2)
ind = 1:ib
add_chord_diagram_legend("bottomleft", ind, all_states[ind], state_col[all_states][ind])
ind = (ib+1):n_states
add_chord_diagram_legend("bottomright", ind, all_states[ind], state_col[all_states][ind])

text(-1.05, 0.5, "group2", adj = c(0, 0.5))
text(-1.05, -0.5, "group1", adj = c(0, 0.5))

suppressPackageStartupMessages(library(gridBase))
suppressPackageStartupMessages(library(ComplexHeatmap))
vps = baseViewports()
pushViewport(vps$inner, vps$figure, vps$plot)

at = round(c(0.5 - abs_max, 0.5, 0.5 + abs_max), digits = 1)
lgd1 = Legend(at = at, labels = at, direction = "horizontal", col_fun = col_fun, title_position = "topcenter",
	title = "Mean methylation", legend_width = unit(4, "cm"))

at = round(c(-abs_max, 0, abs_max), digits = 1)
lgd2 = Legend(at = at, labels = at, direction = "horizontal", col_fun = col_fun2, title_position = "topcenter",
	title = "Mean difference", legend_width = unit(4, "cm"))

pushViewport(viewport(x = unit(0.5, "npc") - unit(3, "cm"), y = unit(-2, "cm"), width = grobWidth(lgd1), 
	height = grobHeight(lgd1), just = c("right", "bottom")))
grid.draw(lgd1)
upViewport()
pushViewport(viewport(x = unit(0.5, "npc") + unit(2, "cm"), y = unit(-2, "cm"), 
	width = grobWidth(lgd2), height = grobHeight(lgd2), just = c("left", "bottom")))
grid.draw(lgd2)
upViewport()
upViewport()
```


<!-- uid=Visualize_chromatin_states_transtitions -->