# Circular layout {#circular-layout}

## Coordinate transformation {#coordinate-transformation}

To map graphics onto the circle, there exist transformations from several
coordinate systems. First, there is a data coordinate system in which ranges
for x-axes and y-axes are the ranges of original data. Second, there is a
polar coordinate system in which these coordinates are mapped onto a circle.
Finally, there is a canvas coordinate system in which graphics are really
drawn on the graphical device (figure \@ref(fig:coordinate-transformation)).
**circlize** first transforms coordinates from data coordinate system to polar
coordinate system and finally transforms into canvas coordinate system. For
users, they only need to imagine that each cell is a normal rectangular
plotting region (data coordinate) in which x-lim and y-lim are ranges of data
in the category respectively. **circlize** knows which cell you are in and
does all the transformations automatically.

```{r coordinate-transformation, echo = FALSE, fig.width = 8, fig.height = 4, fig.cap = "Transformation between different coordinates"}
source("src/intro-03-transformation.R")
```

The final canvas coordinate is in fact an ordinary coordinate in the base R
graphic system with x-range in `(-1, 1)` and y-range in `(-1, 1)` by default.
It should be noted that **the circular plot is always (or mostly except you
want to draw something out of the circle) drawn inside the circle which has
radius of 1 (unit circle), and from outside to inside**.

## Rules for making circular layout {#rules-for-making-circular-layout}

The rule for making circular layout is rather simple. It follows the sequence of
`initialize layout -> create track -> add graphics -> create track -> add graphics - ... -> clear`.
Details are shown in figure \@ref(fig:circlize-order) and as follows:

```{r circlize-order, echo = FALSE, fig.width = 8, fig.height = 7, fig.cap = "Order of drawing circular layout."}
source("src/intro-02-order.R")
```

1. Initialize the layout using `circos.initialize()`. Since circular layout in
   fact visualizes data which is in categories, there must be at least a
   categorical variable. Ranges of x values on each category can be specified
   as a vector or the range itself. See section \@ref(sectors-and- tracks).
2. Create plotting regions for the new track and add graphics. The new track
   is created just inside the previously created one. Only after the creation
   of the track can you add other graphics on it. There are three ways to add
   graphics in cells.

       - After the creation of the track, use low-level graphic function like
         `circos.points()`, `circos.lines()`, ... to add graphics cell by
         cell. It always involves a `for` loop.
       - Use `circis.trackPoints()`, `circos.trackLines()`, ... to add
         graphics through all cells simultaneously. However, it is not
         recommended because it would make code a little confusing to
         understand and also it cannot make complicated graphics.
       - Use `panel.fun` argument in `circos.track()` to add graphics
         immediately after the creation of a certain cell. `panel.fun` needs
         two arguments `x` and `y` which are x-values and y-values that are in
         the current category. This subset operation is applied automatically.
         This is the most recommended way. Section \@ref(panel-fun) gives
         detailed explanation of using `panel.fun` argument.
        
3. Repeat step 2 to add more tracks on the circle unless it reaches the center
   of the circle.
4. Call `circos.clear()` to do the cleanup.

As mentioned above, there are three ways to add graphics on the track. 

1.   Create plotting regions for the whole track first and then add graphics by
     specifying `sector.index`. In the following pseudo code, `x1`, `y1` are
     data points in a given cell, which means you need to do data subsetting by
     yourself.

     In following code, `circos.points()` and `circos.lines()` are used
     separatedly from `circos.track()`, thus, the index for the sector needs to
     be explicitly specified by `sector.index` argument. There is also a
     `track.index` argument for both functions, however, the default value is
     the "current" track index and as the two functions are used just after
     `circos.track()`, the "current" track index is what the two function expect
     and it can be ommited when calling the two functions.

```{r, eval = FALSE}
circos.initialize(factors, xlim)
circos.track(factors, ylim)
for(sector.index in all.sector.index) {
    circos.points(x1, y1, sector.index)
    circos.lines(x2, y2, sector.index)
}
```

2.   Add graphics in a batch mode. In following code, `circos.trackPoints()` and
     `circos.trackLines()` need a categorical variable, a vector of x values and
     a vector of y values. X and y values will be split by the categorical
     variable and sent to corresponding sector to add the graphics. Internally,
     this is done by using `circos.points()` or `circos.lines()` in a `for`
     loop. This way to add graphics would be convenient if users only want to
     add a specific type of graphics (e.g. only points) to the track.

     `circos.trackPoints()` and `circos.trackLines()` need a `track.index` to
     specify which track to add the graphics. Similarly, since they two are
     called just after `circos.track()`, the graphics are added in the newly
     created track right away.

```{r, eval = FALSE}
circos.initialize(factors, xlim)
circos.track(factors, ylim)
circos.trackPoints(factors, x, y)
circos.trackLines(factors, x, y)
```

3.   use a panel function to add self-defined graphics as soon as the cell has
     been created. This is the way recommended and you can find most of the code
     in this documentation uses `panel.fun`. `circos.track()` creates cells one
     by one and after the creation of a cell, `panel.fun` is executed on this cell right away.
     In this case, the "current" sector and "current" track are marked to this cell
     and when using low-level functions, they just add graphics to this cell.

     If you look at following code, you will find the code inside `panel.fun` is as
     natural as using `points()` or `lines()` in the normal R graphic system.
     This is a way to help you think a cell is an "imaginary rectangular plotting region".

```{r, eval = FALSE}
circos.initialize(factors, xlim)
circos.track(factors, all_x, all_y, ylim,
    panel.fun = function(x, y) {
        circos.points(x, y)
        circos.lines(x, y)
})
```

As mentioned above, there are several internal variables keeping tracing of
the current sector and track when applying `circos.track()` and
`circos.update()`. Thus, although functions like `circos.points()`,
`circos.lines()` need to specify the index of sector and track, they will take
the current one by default. As a result, if you draw points, lines, text _et
al_ just after the creation of the track or cell, you do not need to set the
sector index and the track index explicitly and it will be added in the most
recently created cell.

## Sectors and tracks {#sectors-and-tracks}

A circular layout is composed of sectors and tracks. As illustrated in figure
\@ref(fig:circlize-coordinate), the red circle is one track and the blue represent
one sector. The intersection of a sector and a track is called a cell which
can be thought as an imaginary plotting region for data points. In this
section, we introduce how to set data ranges on x and y directions in cells.

```{r circlize-coordinate, echo = FALSE, fig.width = 4, fig.height = 4, fig.cap = "Sectors and tracks in circular layout."}
source("src/intro-04-coordinate.R")
```

Sectors are first allocated on the circle by `circos.initialize()`. There must
be a categorical variable (say `factors`) that on the circle, each sector
corresponds to one category. The width of sectors are proportional to the data
range in sectors on x direction. The data range can be specified as a numeric
vector `x` which has same length as `factors`, then `x` is split by `factors`
and data ranges are calculated for each sector internally.

Data ranges can also be specified directly by `xlim` argument. The valid value
for `xilm` is a two-column matrix with same number of rows as number of
sectors that each row in `xlim` corresponds to one sector. If `xlim` has row
names which already cover sector names, row order of `xlim` is automatically
adjusted.

```{r, eval = FALSE}
circos.initialize(factors, x = x)
circos.initialize(factors, xlim = xlim)
```

After the creation of the layout, you may not see anything drawn or only an
empty graphical device is opened. That is because no track has been created
yet, however, the layout has already been recorded internally.

There are something very important that should be noted in the initialization
step. In this step, not only the width of each sector is assigned, but also
the order of sectors on the circle is determined. **Order of the sectors are
determined by the order of levels of the input factor**. If the value for
`factors` is not a factor, the order of sectors is `unique(factors)`. Thus, if
you want to change the order of sectors, you can just change of the level of
{\tt factors} variable. The following code generates plots with different
sector orders (Figure \@ref(fig:circlize-factor):

```{r, eval = FALSE}
fa = c("d", "f", "e", "c", "g", "b", "a")
f1 = factor(fa)
circos.initialize(factors = f1, xlim = c(0, 1))
f2 = factor(fa, levels = fa)
circos.initialize(factors = f2, xlim = c(0, 1))
```

```{r circlize-factor, echo = FALSE, fig.width = 8, fig.height = 4, fig.cap = "Different sector orders."}
source("src/intro-05-factor.R")
```

**In different tracks, cells in the same sector share the same data range on
x-axes.** Then, for each track, we only need to specify the data range on y
direction for cells. Similar as `circos.initialize()`, `circos.track()` also
receives either `y` or `ylim` argument to specify the range of y-values. Since
all cells in a same track shares a same y range, `ylim` is just a vector of
length two if it is specified.

In Section \@ref(panel-fun), we will introduce how `x` and `y` are sent to each cell
and how the graphics are added.

```{r, eval = FALSE}
circos.track(factors, y = y)
circos.track(factors, ylim = ylim)
```

In the track creation step, since all sectors have already been allocated in
the circle, if `factors` argument is not set, `circos.track()` would create
plotting regions for all available sectors. Also, levels of `factors` do
not need to be specified explicitly because the order of sectors has already
be determined in the initialization step. If users only create cells for a part
of sectors in the track (not all sectors), in fact, cells in remaining
unspecified sectors are created as well, but with no borders (pretending they
are not created).

```{r, eval = FALSE}
# assume `factors` only cover subset of sectors
# you will only see cells covered in `factors`
circos.track(factors, y = y)
# all cells are covered
circos.track(ylim = ranges(y))
```

Cells are basic units in the circle and are independent with each other. After
the creation of cells, they have self-contained meta values of x-lim and y-lim
(data range measured in data coordinate). So if you are adding graphics in one
cell, you do not need to consider things outside the cell and also you do not
need to consider you are in the circle. Just pretending it is rectangle area.

```{r circlize-direction, echo = FALSE, fig.width = 8, fig.height = 4, fig.cap = "Sector directions."}
source("src/intro-07-direction.R")
```

## Graphic parameters {#graphic-parameters}

Some basic parameters for the circular layout can be set by `circos.par()`.
The parameters are listed as follows. Note some parameters can only be
modified before the initialization of the circular layout.

- `start.degree`: The starting degree where the first sector is put. Note this
  degree is measured in the standard polar coordinate which means it is always
  reverse clockwise. See figure \@ref(fig:circlize-direction).
- `gap.degree`: Gap between two neighbour sectors. It can be a single value
  which means all gaps share same degree, or a vector which has same number as
  sectors. ** Note the first gap is after the first sector.** See figure
  \@ref(fig:circlize-direction) and figure \@ref(fig:circlize_region).
- `gap.after`: Same as `gap.degree`, but more understandable. Modifying values
  of `gap.after` will also modify `gap.degree` and vice versa.
- `track.margin`: Like `margin` in Cascading Style Sheets (CSS), it is the
  blank area out of the plotting region, also outside of the borders. Since
  left and right margin are controlled by `gap.degree`, only bottom and top
  margin need to be set. The value for `track.margin` is the percentage to the
  radius of the unit circle. The value can also be set by `uh()` function with
  absolute units. See figure \ref@(fig:circlize-region).
- `cell.padding`: Padding of the cell. Like `padding` in Cascading Style
  Sheets (CSS), it is the blank area around the plotting regions, but within
  the borders. The parameter has four values, which control the bottom, left,
  top and right padding respectively. The first and the third padding values
  are the percentages to the radius of the unit circle, and the second and
  fourth values are the degrees. The first and the third value can be set by
  `uh()` with absolute units. See figure \@ref{fig:circlize-region}.
- `unit.circle.segments`: Since curves are simulated by a series of straight
  lines, this parameter controls the amount of segments to represent a curve.
  The minimal length of the line segment is the length of the unit circle
  ($2\pi$) divided by `unit.circle.segments`. More segments means better
  approximation for the curves, while generate larger file size if figures are
  in PDF format. See explanantion in Section \@ref(lines).
- `track.height`: The default height of tracks. It is the percentage to the
  radius of the unit circle. The height includes the top and bottom cell
  paddings but not the margins. The value can be set by `uh()` with absolute
  units.
- `points.overflow.warning`: Since each cell is in fact not a real plotting
  region but only an ordinary rectangle (or more precisely, rectangle-like),
  it does not remove points that are plotted outside of the region. So if some
  points are out of the plotting region, by default, the package would
  continue drawing the points and print warnings. But in some circumstances,
  draw something out of the plotting region is useful, such as adding some
  text annotations. Set this value to `FALSE` to turn off the warnings.
- `canvas.xlim`: The coordinate for the canvas in x direction. **circlize** is
  forced to put everything inside the unit circle, so `xlim` and `ylim` for
  the canvas would be `c(-1, 1)` by default. However, you can set it to a more
  broad interval if you want to draw other things out of the circle. By
  choosing proper `canvas.xlim` and `canvas.ylim`, you can only draw part of
  the circle. E.g. setting `canvas.xlim` to `c(0, 1)` and `canvas.ylim` to
  `c(0, 1)` would only draw circle in the region of `(0, pi/2)`.
- `canvas.ylim`: The coordinate for the canvas in y direction.
- `clock.wise`: The order for drawing sectors. Default is `TRUE` which means
  clockwise (figure \@ref(fig:circlize-direction). **But note that inside each
  cell, the direction of x-axis is always clockwise and direction of y-axis is
  always from inside to outside in the circle.**

```{r circlize-region, echo = FALSE, fig.width = 4, fig.height = 4, fig.cap = "Regions in a cell"}
source("src/intro-06-region.R")
```

Default values for graphic parameters are in following table.

-------------------------- ----------------------------
`start.degree`             `0` 
`gap.degree`/`gap.after`   `1`
`track.margin`             `c(0.01, 0.01)`
`cell.padding`             `c(0.02, 1.00, 0.02, 1.00)`
`unit.circle.segments`     `500`
`track.height`             `0.2`
`points.overflow.warning`  `TRUE`
`canvas.xlim`              `c(-1, 1)`
`canvas.ylim`              `c(-1, 1)`
`clock.wise`               `TRUE` 
-------------------------- ---------------------------

Parameters related to the allocation of sectors cannot be changed after the
initialization of the circular layout. Thus, `start.degree`,
`gap.degree`/`gap.after`, `canvas.xlim`, `canvas.ylim` and `clock.wise` can
only be modified before `circos.initialize()`. The second and the fourth
values of `cell.padding` (left and right paddings) can not be modified neither
(or will be ignored).

## Create plotting regions {#create-plotting-regions}

As described above, only after creating the plotting region can you add low-
level graphics on it. The minimal set of arguments for this function is to set
either `y` or `ylim` which assigns range of y-values for the track.
`circos.track()` creates tracks for all sectors although in some case only
parts of them are visible.

If `factors` is not specified, all cells in the track will be created with the
same settings. If `factors`, `x` and `y` are set, they need to be vectors with
the same length. Proper values of `x` and `y` that correspond to current cell
will be passed to `panel.fun` by subsetting `factors` internally. Section
\@ref(panel-fun) explains the usage of `panel.fun`.

Graphic arguments such as `bg.border` and `bg.col` can either be a scalar or a
vector. If it is a vector, the length must be equal to the number of sectors.
Thus, you can create plot regions with different styles of borders and
background colors.

If you are confused with the `factors` orders, you can also customize the
borders and background colors inside `panel.fun`.
`get.cell.meta.data("cell.xlim")` and `get.cell.meta.data("cell.ylim")` give
you dimensions of the plotting region and you can customize plot regions
directly by `circos.rect()`.

## Update plotting regions {#update-plotting-regions}

`circos.track()` creates new tracks, however, if `track.index` argument is set
to a track which already exists, `circos.track()` actually **re-creates** this
track. In this case, settings related to the positions of the track such as
the height of the track can not be modified.

```{r, eval = FALSE}
circos.track(factors, ylim = c(0, 1), track.index = 1, ...)
```

For a single cell, `circos.update()` can be used to erase all graphics that
have been already added in the cell. However, the data coordinate in the cell
still keeps unchanged.

```{r, eval = FALSE}
circos.update(sector.index, track.index)
circos.points(x, y, sector.index, track.index)
```

## `panel.fun` argument {#panel-fun}

`panel.fun` argument in `circos.track()` is extremely useful to apply plotting
as soon as the cell has been created. This self-defined function needs two
arguments `x` and `y` which are data points that belong to this cell. The
value for `x` and `y` are automatically extracted from `x` and `y` in
`circos.track()` according to the category defined in `factors`. In the
following example, inside `panel.fun`, in sector `a`, `x` is `1:3` and in
sector `b`, `x` is `4:5`. If `x` or `y` in `circos.track()` is `NULL`, then
`x` or `y` inside `panel.fun` is also `NULL`.

```{r, eval = FALSE}
factors = c("a", "a", "a", "b", "b")
x = 1:5
y = 5:1
circos.track(factors = factors, x = x, y = y,
    panel.fun = function(x, y) {
        circos.points(x, y)
})
```

In `panel.fun`, one thing important is that if you use any low-level graphic
functions, you don't need to specify `sector.index` and `track.index`
explicitly. Remember that when applying `circos.track()`, cells in the track
are created one after one. When a cell is created, **circlize** would set the
sector index and track index of the cell as the 'current' index for the sector
and track. When the cell is created, `panel.fun` is executed immediately.
Without specifying `sector.index` and `track.index`, the 'current' ones are
used and that's exactly what you need.

The advantage of `panel.fun` is that it makes you feel you are using graphic
functions in the base graphic engine (You can see it is almost the same of
using `circos.points(x, y)` and `points(x, y)`). It will be much easier for
users to understand and customize new graphics.

Inside `panel.fun`, information of the 'current' cell can be obtained through
`get.cell.meta.data()`. Also this function takes the 'current' sector and
'current' track by default.

```{r, eval = FALSE}
get.cell.meta.data(name)
get.cell.meta.data(name, sector.index, track.index)
```

Information that can be extracted by `get.cell.meta.data()` are:

- `sector.index`: The name (label) for the sector.
- `sector.numeric.index`: Numeric index for the sector.
- `track.index`: Numeric index for the track.
- `xlim`: Minimal and maximal values on the x-axis.
- `ylim`: Minimal and maximal values on the y-axis.
- `xcenter`: mean of `xlim`.
- `ycenter`: mean of `ylim`.
- `xrange`: defined as `xlim[2] - xlim[1]`.
- `yrange`: defined as `ylim[2] - ylim[1]`.
- `cell.xlim`: Minimal and maximal values on the x-axis extended by cell
  paddings.
- `cell.ylim`: Minimal and maximal values on the y-axis extended by cell
  paddings.
- `xplot`: Degree of right and left borders in the plotting region. The first
  element corresponds to the start point of values on x-axis and the second element corresponds to the end point of values
  on x-axis Since x-axis in data coordinate in cells are
  always clockwise, `xplot[1]` is larger than `xplot[2]`.
- `yplot`: Radius of bottom and top radius in the plotting region.
- `cell.start.degree`: Same as `xplot[1]`.
- `cell.end.degree`: Same as `xplot[2]`.
- `cell.bottom.radius`: Same as `yplot[1]`.
- `cell.top.radius`: Same as `yplot[2]`.
- `track.margin`: Margins of the cell.
- `cell.padding`: Paddings of the cell.

Following example code uses `get.cell.meta.data()` to add sector index in the
center of each cell.

```{r, eval = FALSE}
circos.track(ylim = ylim, panel.fun = function(x, y) {
    sector.index = get.cell.meta.data("sector.index")
    xcenter = get.cell.meta.data("xcenter")
    ycenter = get.cell.meta.data("ycenter")
    circos.text(xcenter, ycenter, sector.index)
})
```

`get.cell.meta.data()` can also be used outside `panel.fun`, but you need to
explictly specify `sector.index` and `track.index` arguments or you need to be
aware of which is your 'current' track and sector.

There is a companion variable `CELL_META` which is identical to
`get.cell.meta.data()` to get cell meta information, but easier and shorter to
write. Actually, the value of `CELL_META` itself is meaningless, but e.g.
`CELL_META$sector.index` is automatically redirected to
`get.cell.meta.data("sector.index")`. Following code rewrites above example
code with `CELL_META`.

```{r, eval = FALSE}
circos.track(ylim = ylim, panel.fun = function(x, y) {
    circos.text(CELL_META$xcenter, CELL_META$ycenter, 
        CELL_META$sector.index)
})
```

Please note `CELL_META` only extracts information for the "current" cell, thus,
it is only recommended to use in `panel.fun`.

